---
title: "_`R`_: Module _`1`_ Basics for reproducible data analysis"
description: "Learn the basics of using R for statistical data analysis"
author:
  - name: "Anoob Prakash"
    email: "prakas89@purdue.edu"
    orcid: "0000-0002-7207-934X"
    url: "https://www.linkedin.com/in/anoob-prakash/"
date: "09/25/2026"
date-modified: "10/08/2026"
categories:
  - hpc
  - r


format:
  html:
    toc: true
    toc-location: left
    toc-title: "On this page"
    toc-depth: 3
    page-layout: full
    lightbox: true
    code-fold: false
    code-overflow: wrap
    code-copy: true
    code-tools: false
    code-block-border-left: "#583C88"
    code-block-border-left-size: 5px
    code-block-border-left-style: solid
---




## Goal


Learn core R syntax and data-handling skills by analyzing the red-spruce fitness-trait dataset you already downloaded in [*Bashing the cluster*](/notebook/posts/bashing-the-cluster/index.qmd), and finish with a first reproducible R script whose outputs land in `results/`.


This module picks up exactly where the cluster exercise left off: `data/raw/` holds the raw file, `src/` will hold your R scripts, and `results/` will hold everything R produces.


#### Prerequisites
::: {.column-body}
- Completed the cluster exercise 
- R (>= 4.2) installed; RStudio/Positron recommended but not required
- The dataset available at `data/raw/red_spruce_fitness_traits.txt`


If you're working locally instead of on the cluster, grab the same file directly from R:


```{r}
#| eval: false
# Create directory
dir.create("data/raw", 
recursive = TRUE,      # create the last element of the path- here it will be "raw"
showWarnings = FALSE)


download.file(
"https://raw.githubusercontent.com/anoobvinu07/Genomic_assisted_selection/master/data/FitnessTraits_GeneticParameters_RedSpruce.txt",
      destfile = "data/raw/red_spruce_fitness_traits.txt")
```  


This is work on my system, probably not on yours. Lets use a package called `here` to make sure that code is reproducibile on your system as well.
```{r}
#| eval: false
# Create directory
dir.create(here::here("data/raw"), 
recursive = TRUE,      # create the last element of the path- here it will be "raw"
showWarnings = FALSE)


download.file(
"https://raw.githubusercontent.com/anoobvinu07/Genomic_assisted_selection/master/data/FitnessTraits_GeneticParameters_RedSpruce.txt",
      destfile = here::here("data/raw/red_spruce_fitness_traits.txt"))
``` 



::: {.callout-tip}
# Code reproducibility check


I am using the `here` package to make this script reproducible. This makes sure that irrespective of of which system this code is run, it will work on that system. Read more about `here` [here](https://github.com/jennybc/here_here#readme).


Instead of calling the package specifically at the top of the script with a `library(here)` or `require(here)`, I am calling them as a require it. That is done with the syntax `packageName::packageFunction`. This is useful when you have a lot of packages at the top, but you want to keep track of where a specific function is being called from. There are other reasons as well for following this syntax, for example when you need to call a function that is part of a package not part of the base R. Have you encounetered such situations while coding?
:::

::: {.callout-note icon="false" appearance="default" collapse="true"}
## Logic check: what does `here::here()` actually do?

`here::here()` answers two questions at once:

1. **Where is the project root?** It walks up the directory tree from your
   working directory until it finds a marker file — a `.Rproj` file, a `.git`
   folder, or a few others. That marker's directory *is* the project root.
2. **How do I build a path inside it?** Any arguments you pass become folders
   in order: `here::here("data", "raw")` returns
   `<project-root>/data/raw` with the right separator for your operating
   system (`/` on macOS/Linux, `\` on Windows).

Two consequences worth remembering:

- The same line, `here::here("data/raw/file.txt")`, produces the *absolute*
  path on your laptop, on the cluster, and on a collaborator's Windows machine
  — you never type an absolute path again.
- If you open R in the *wrong* directory (above or beside the project root),
  `here::here()` finds no marker and silently falls back to your current
  working directory — which is why the `getwd()` / `here::here()` check below
  is worth running at the start of every session.
:::

Recreate the rest of the project layout if you don't already have it:
```{r}
dir.create(here::here("results"), showWarnings = FALSE)
dir.create(here::here("src"), showWarnings = FALSE)
```




::: {.callout-note}


What is the output when you type in the following commands in your system? What does it mean?


```{r}
#| eval: false
getwd()


here::here()
```
:::

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Answer

- `getwd()` prints your **current working directory** — the folder R treats as
  the origin for every relative path you type. If you `read.delim("x.txt")`,
  R looks for `x.txt` there. On the cluster it might be
  `/home/youruser/projects/cluster-exercise`; on a laptop,
  `/Users/you/Documents/cluster-exercise`.
- `here::here()` called with no arguments prints the **project root** it
  detected. On a healthy setup, `getwd()` and `here::here()` usually print the
  same folder — and if they don't, you've opened R somewhere inside the project
  (say, in `src/`), and `here` will still resolve paths correctly while bare
  relative paths would break. That difference is exactly the bug `here`
  prevents.
:::




::: {.callout-important collapse="true"}
## Project directory setup
**This excercise assumes the project folder follows the following structure:** 


<pre style="color: #F8F9FA; background-color: #22252A; padding: 1rem; border-radius: 6px;"><code>cluster-exercise/
 ├── data/
 │   └── raw/       # downloaded, unchanged source data
 ├── logs/          # Slurm standard output and error logs
 ├── results/       # analysis products created by R
 └── src/           # R and Slurm scripts</code></pre>
:::



:::


## Learning objectives


By the end of this module you should be able to:


- Start R/RStudio with a reproducible working directory (no hard-coded absolute paths)
- Describe R's core data types and structures: vectors, factors, data frames, lists
- Import and inspect a tab-delimited file
- Subset, filter, and summarize data with base R
- Produce basic plots (histogram, boxplot, scatterplot)
- Save tables and plots into `results/`
- Turn interactive code into a script you can run non-interactively with `Rscript`


#### R quick reference


| Function/operator | Purpose | Example |
|---|---|---|
| `c()` | Combine values into a vector | `c(1, 2, 3)` |
| `<-` | Assignment | `x <- 5` |
| `class()` | Report the type of an object | `class(traits$Height)` |
| `is.na()` | Flag missing values | `sum(is.na(traits$Height))` |
| `na.rm = TRUE` | Skip NAs inside a computation | `mean(x, na.rm = TRUE)` |
| `read.delim()` | Read a tab-delimited file | `read.delim("data/raw/file.txt")` |
| `str()` | Show structure/types of an object | `str(df)` |
| `dim()`, `nrow()`, `ncol()` | Dimensions of a data frame | `dim(df)` |
| `head()` / `tail()` | First/last rows | `head(df, 5)` |
| `summary()` | Quick numeric/factor summary | `summary(df$Height)` |
| `df$col` | Access a column by name | `df$Fitness` |
| `df[rows, cols]` | Subset by position/condition | `df[df$Region == "E", ]` |
| `table()` | Count occurrences | `table(df$Population)` |
| `tapply()` / `aggregate()` | Grouped summaries | `tapply(df$Height, df$Population, mean)` |
| `function(...) {}` | Define a function | `f <- function(x) mean(x, na.rm = TRUE)` |
| `for (i in seq)` | Loop | `for (p in unique(df$Population)) {...}` |
| `hist()`, `boxplot()`, `plot()` | Base R plotting | `hist(df$Height)` |
| `write.csv()` | Save a table | `write.csv(x, "results/out.csv")` |

## Setting up your R project


On the cluster, load R and start it from inside `cluster-exercise/`:


```{bash}
#| eval: false


cd ~/projects/cluster-exercise


module biocontainers/default
module load r          # module name may differ on your cluster; check `module avail r`


R                      # start the R session
```



Inside R, always confirm where you are before reading or writing anything:


```{r}
getwd()


#or


here::here() # when using the here package to set the paths
```


If you're using RStudio, create an RStudio Project (`File > New Project`) rooted at `cluster-exercise/` — this keeps every path relative and portable, exactly like the `data/raw/...` paths from the bash module.


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Logic check: why does the session start matter?

A script that begins by assuming "I am in the right folder" works today and
breaks the day you launch R from one directory higher, or copy the project to
the cluster. The 10 seconds you spend checking `getwd()` up front converts a
confusing `cannot open file 'data/raw/...'` error into a one-line fix
(`setwd()` to the right place, or better, restart R in the project root).

Rule of thumb: **check the location first, then read files.** Every
"file not found" error is either a wrong working directory, a typo, or a file
that genuinely isn't there — and the first cause is by far the most common.
:::


## R data types and structures


```{r}
# Vectors: one-dimensional, single type
heights <- c(27.3, 31.1, 29.6, 32.0)
pops    <- c("AB", "AB", "CC", "CC")


# Factors: categorical vectors with fixed levels
region <- factor(c("E", "E", "C", "C"))
levels(region)


# Data frames: a table — columns can be different types
df_demo <- data.frame(pop = pops, height = heights)
str(df_demo)


# Lists: can hold mixed, differently-shaped objects
result <- list(mean_height = mean(heights), populations = unique(pops))
result$mean_height
```


**Key idea:** almost everything you do in this module is a data frame — one row per tree, one column per variable — with vectors as its columns.


### The four structures, one at a time

::: {.panel-tabset}

#### Vectors

A vector is R's atomic unit: **one dimension, one type**. Everything you
compute with — a column of heights, a column of region labels — is a vector.

```{r}
heights            # print it
class(heights)     # what type is it? "numeric"
length(heights)    # how many elements?
class(pops)        # "character"
```

A vector's type is decided by the *most flexible* element inside it — this is
called **implicit coercion**, and it is the source of many silent surprises:

```{r}
c(1, 2, 3)        # numeric: 1 2 3
c(1, "a")         # character: "1" "a" — the number was demoted to text!
c(TRUE, 1)        # numeric: 1 1      — TRUE became 1
```

The hierarchy is `logical → numeric → character`: mixing types in one `c()`
promotes everything to the highest type present. Nothing errors, nothing warns —
your `1` just silently becomes `"1"` and later comparisons like
`x == 1` stop matching.

#### Factors

A factor is a categorical vector with a fixed, ordered set of **levels**:

```{r}
region
as.integer(region)   # stored internally as 2 2 1 1 (E=2, C=1: levels are alphabetical)
table(region)        # counts per level
```

Two things make factors different from plain text:

1. **Memory and clarity:** a column with 326 `"C"`, `"E"`, `"M"` strings stores
   the full text repeatedly; a factor stores small integers plus the level
   table once.
2. **Fixed classes:** `levels()` defines which categories *exist*, even if
   they have zero observations. This matters for counting — as you'll see with
   genotypes in population genetics module, `table()` on a plain vector drops categories that
   don't occur, while `table(factor(x, levels = ...))` keeps them with a
   count of zero.

Plotting and modeling functions (`boxplot`, `lm`) use levels to decide the
order of groups on the axis, which is why the idea check asks about
`boxplot(Height ~ Region, ...)`.

#### Data frames

A data frame is a **list of equal-length vectors** displayed as a table — which
is why columns can have different types but every column has the same number
of rows. It is the structure you will use 95% of the time:

```{r}
df_demo
nrow(df_demo)   # rows = observations
ncol(df_demo)   # columns = variables
```

One row = one observation (one tree); one column = one variable (its height,
its region). This "tidy" layout is what `aggregate()`, `boxplot(Height ~ Region)`,
and almost every R modeling function expect.

#### Lists

A list is a **container with no constraints**: elements can be different types
*and* different lengths — even other lists. They are how functions bundle
multiple outputs:

```{r}
result
names(result)    # each slot has a name
result$mean_height
```

You will mostly *read* lists (model objects, `str()` output) before you write
them, but recognizing `[[1]]` vs `$name` access patterns pays off immediately
when you start calling `summary()` on fitted models.

:::


### Missing values: `NA`, the silent saboteur


Real datasets have missing data, and this one is no exception — 13 trees are missing
a `Height` measurement. R represents a missing data with `NA`, and the rule you must
internalize is:

> **Almost any computation that touches an `NA` returns `NA`.**

```{r}
x <- c(27.3, 31.1, NA, 32.0)

mean(x)              # NA — R refuses to guess
mean(x, na.rm = TRUE) # 30.13 — "remove NAs, then compute"

is.na(x)              # FALSE FALSE  TRUE FALSE — a logical flag per element
sum(is.na(x))         # 1 — counting NAs by summing TRUEs
sum(!is.na(x))        # 3 — counting non-missing values
```

Why does `sum(is.na(x))` work? `is.na()` returns a logical vector, and in
arithmetic `TRUE` counts as 1 and `FALSE` as 0 — so summing counts the
`TRUE`s. That idiom (`sum(is.na(col))`) is the standard first move when
inspecting any new dataset.

`NA` is *contagious* in subsetting too — see the indexing section below.


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Logic check: predict the output

Before running each line, write down what you expect. Then check.

```{r}
c(1, 2, 3) > 2
mean(c(1, NA, 3))
sum(c(TRUE, TRUE, FALSE))
class(c(TRUE, 2))
table(c("A", "A", "B", NA))
```
:::

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Answers

- `c(1, 2, 3) > 2` → `FALSE FALSE TRUE`. Comparisons are vectorized:
  the operation applies element-wise, and the result is a logical vector.
  This vectorized comparison is the engine behind filtering rows.
- `mean(c(1, NA, 3))` → `NA`. No `na.rm`, no answer. Add `na.rm = TRUE`
  to get `2`.
- `sum(c(TRUE, TRUE, FALSE))` → `2`. Booleans are 1/0 under arithmetic —
  the counting trick from the NA section.
- `class(c(TRUE, 2))` → `"numeric"`. Coercion promotes `TRUE` to `1`,
  because mixing logical and numeric yields numeric.
- `table(c("A", "A", "B", NA))` → counts for `A` (2) and `B` (1) only, and
  **the NA is silently excluded** from the printed table. When checking a
  dataset, `sum(is.na(x))` is more trustworthy than eyeballing `table()`.
:::


### Importing and inspecting the dataset


```{r}
traits <- read.delim(here::here("data/raw/red_spruce_fitness_traits.txt"),
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)


dim(traits)          # rows x columns
names(traits)        # column names
str(traits)          # types per column
head(traits, 5)      # first 5 rows
summary(traits$Height)
```


You should see 19 columns including identifiers (`Family`, `Population`, `Tree`, `Location`, `Region`), coordinates and environment (`Latitude`, `Longitude`, `Elevation`), fitness traits (`SeedWeight`, `Germination`, `Survival`, `Height`, `Fitness`), and genetic parameters (`PC1`, `PC2`, `Family_Homozygosity`, `Population_Homozygosity`, `Genetic_Diversity`, `Genetic_Load`) — we'll use that last group in the Population genetics foundation module.


**Don't assume a column's type from its name** — Always check with `str()` before analyzing.

::: {.callout-note icon="false" appearance="default" collapse="true"}
## Code breakdown: what each inspection function told you

- `dim(traits)` → `326 19`: 326 rows (trees) and 19 columns (variables).
  Confirm the row count matches your expectation from the raw file — a
  mismatch usually means a wrong `sep` or `header` argument.
- `names(traits)` → the 19 column names. You will type these constantly;
  check spelling here first when `object not found` errors appear later.
- `str(traits)` → the full type map: identifiers are `chr` (text), `Tree`
  and `Elevation` are `int` (integers), everything numeric with decimals is
  `num`. The three-letter prefixes mean: `chr` = character, `int` = integer,
  `num` = numeric (double). `int` columns *are* numeric for computation —
  the distinction only matters for storage and factor conversion.
- `head(traits, 5)` → the first 5 rows *as they sit in the file* — useful for
  spotting parsing disasters (values in the wrong column, text where numbers
  should be).
- `summary(traits$Height)` → Min 19.39, Median 29.94, Mean 30.03, Max 47.56 —
  **and 13 `NA`s**. The dataset genuinely has 13 trees with no height
  measurement. This single line is why the NA section above exists: without
  `na.rm = TRUE`, every mean, sd, and correlation on this column is `NA`.
:::

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: reading a new file safely

The routine to run on *any* new tabular file, before you analyze anything:

```{r}
#| eval: false
# 1. How big is it? Does the shape match the source?
dim(traits)

# 2. What are the columns called, and what type did R guess?
str(traits)

# 3. Does the first rows look like the actual file (compare to the raw text)?
head(traits)

# 4. Where are the holes?
colSums(is.na(traits))

# 5. Do the categorical columns have the levels I expect?
table(traits$Region)
```

For this dataset step 4 reports: `SeedWeight` 12, `Germination` 12,
`Survival` 1, `Height` 13, `Fitness` 12, `Elevation` 3 — everything else is
complete. Knowing this *before* computing anything is what separates a
smooth analysis from an hour of "why is my mean `NA`?"
:::


### Indexing and subsetting


```{r}
# A single column
traits$Fitness |> head() # |> is called the pipe command, which lets you run the additional commands (head()) at the end 


# Rows meeting a condition
edge <- traits[traits$Region == "E", ]
nrow(edge)


# Multiple conditions
healthy_edge <- traits[traits$Region == "E" & traits$Survival > 0.8, ]


# subset() reads more like a sentence
subset(traits, Region == "E" & Height > 35, select = c(Population, Height, Fitness))
```


#### How the bracket syntax works

`df[rows, cols]` — the comma splits rows from columns, and an empty slot means
"everything":

| Code | Meaning |
|---|---|
| `df[1, ]` | row 1, all columns |
| `df[, 2]` | all rows, column 2 |
| `df[1:5, c("Height", "Fitness")]` | rows 1–5, two named columns |
| `df[df$Region == "E", ]` | every row where the condition is TRUE, all columns |
| `df[, -1]` | all rows, drop column 1 |

The filter `df[condition, ]` works because the condition is a **logical vector
with one TRUE/FALSE per row**, and bracket-indexing keeps exactly the rows
whose position is `TRUE`. That is the vectorized comparison from the logic
check, doing real work.

Three beginner traps:

1. **`==` vs `=`.** `==` compares ("is Region equal to E?"); `=` assigns or
   names arguments. `traits[traits$Region = "E", ]` is a syntax error.
2. **`&` vs `|`.** `&` is AND (both conditions), `|` is OR (either). For
   filter vectors use the single forms — `&&`/`||` only compare single values.
3. **NA rows vanish.** The one tree with missing `Survival` makes
   `traits$Survival > 0.8` evaluate to `NA` for that row, and `[` drops `NA`
   rows silently. Usually harmless (that row can't satisfy the condition
   anyway) — but if you need to keep or count them explicitly, use
   `which()` or `complete.cases()`:

```{r}
# Explicit about missing data: keep only rows with no NA anywhere
traits_complete <- traits[complete.cases(traits), ]
nrow(traits_complete)   # rows surviving the NA filter
```


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Logic check: what did the subsetting code return?

- `edge` keeps **101 rows** — the number of trees from Region "E". Nothing
  about the filter changed column count: `ncol(edge)` is still 19.
- `healthy_edge` applies **two** conditions at once with `&`; only rows that
  are Region "E" *and* have `Survival` above 0.8 survive.
- `subset(..., select = c(...))` does both jobs in one call: filter rows and
  pick columns — its output has only 3 columns. Note it's a convenient
  interactive shortcut; the bracket syntax is what scripts and functions
  reliably use.
:::


### Summarizing data


```{r}
# Counts
table(traits$Population) # number of inds per population
table(traits$Region)     # number of inds per region


# Grouped means with tapply()
tapply(traits$Height, traits$Population, mean, na.rm = TRUE)


# Grouped summary with aggregate()
aggregate(cbind(Height, Fitness) ~ Population, data = traits, FUN = mean) |> head()
```


#### What each summarizer is for

| Function | Question it answers | Key arguments |
|---|---|---|
| `table()` | How many observations per category? | the categorical vector |
| `tapply()` | Apply one function to one column, split by one group | (values, groups, function) |
| `aggregate()` | Same, but multiple columns and multiple groups at once | formula like `cbind(a, b) ~ group` |

The formula syntax (`Height ~ Population`) reads right-to-left: "height,
explained by population" — split Height by Population. `cbind()` glues several
response columns together so one call summarizes all of them.

Notice the `na.rm = TRUE` inside `tapply()`: it is passed through to `mean()`.
Without it, every population containing one of the 13 missing heights would
report `NA` — and with this dataset, that would be most of them. When a
grouped summary comes back full of `NA`s, this is the first thing to check.

```{r}
# The formula version of aggregate with two grouping factors
aggregate(Height ~ Region, data = traits, FUN = mean)
```


### Basic plotting


:::{.panel-tabset}
#### Default Histogram
```{r}
hist(traits$SeedWeight, main = "Seed weight distribution", 
        xlab = "Seed weight")
```


#### Breaks = 30
```{r}
hist(traits$SeedWeight, main = "Seed weight distribution", 
        xlab = "Seed weight", breaks = 30)
```



#### Breaks = 50
```{r}
hist(traits$SeedWeight, main = "Seed weight distribution", 
        xlab = "Seed weight", breaks = 50)
```
:::

::: {.callout-note icon="false" appearance="default" collapse="true"}
## Logic check: what does `breaks` actually do?

A histogram first bins the data — it cuts the value range into intervals
("bins") and counts observations per bin. `breaks` is a *suggestion* for how
many bins to use:

- **Default** (~10 bins): the shape is blocky; broad features only.
- **breaks = 30**: medium detail — real structure appears (this seed-weight
  distribution is right-skewed with a long thin tail).
- **breaks = 50**: fine detail, but each bin holds fewer trees, so the counts
  get noisier and the shape gets jittery.

The trade-off is **bias vs. variance**: too few bins smooth over real features;
too many bins amplify sampling noise. There is no single right answer — the
point of the tabset is that you *look* at several and pick one that shows the
structure without chasing noise. The x-axis is the variable's value; the
y-axis (default `freq = TRUE`) is the raw count per bin.
:::

```{r}
boxplot(Height ~ Region, data = traits, 
        main = "Height by region", 
        ylab = "Height (cm)")
```

::: {.callout-note icon="false" appearance="default" collapse="true"}
## Logic check: how to read this boxplot

For each region the box shows five numbers:

- **Bold center line** — the median: C ≈ 29.8, E ≈ 29.0, M ≈ 33.6.
- **Box edges** — the 1st and 3rd quartiles; the box spans the middle 50% of
  trees (the interquartile range, IQR).
- **Whiskers** — the furthest values within 1.5 × IQR of the box.
- **Points beyond whiskers** — individual outlier trees.

What the plot says: all three regions overlap heavily (heights between
regions are not dramatically different), but region M sits visibly higher —
its median exceeds the upper quartile of E. The medians differ even though
the distributions overlap, which is precisely the situation where a
*comparison of groups* (boxplot) is more informative than the overall
histogram. Note `summary()` gives you these same five numbers numerically —
the boxplot is `summary()` drawn.
:::

```{r}
plot(traits$Elevation, traits$Fitness,
     xlab = "Elevation (m)", ylab = "Fitness",
     main = "Fitness vs elevation")
```


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Logic check: reading a scatterplot

The cloud trends **downhill**: higher-elevation trees tend to have lower
fitness. The correlation (computed in the exercises) is about **−0.36** — a
moderate negative association, not a tight line.

Two cautions before interpreting:

1. **Association is not causation.** Elevation here stands in for a whole bundle of correlated variables (temperature, growing season, soil). The plot cannot tell you *which* one matters.
2. **Correlation ignores structure.** The three regions occupy different elevation bands. A trend across regions might be a region effect (genetic origin) rather than an elevation effect — the confounding theme population genetics module tackles head-on with PCA before genotype–environment association.

Adding a fitted line summarizes the trend in one stroke:

```{r}
#| eval: false
plot(traits$Elevation, traits$Fitness,
     xlab = "Elevation (m)", ylab = "Fitness")
abline(lm(Fitness ~ Elevation, data = traits), col = "red", lwd = 2)
```

`abline()` draws a straight line from a model fit; `lm()` fits it. The
`col`/`lwd` options style the line so it stands out from the points.
:::


### Saving outputs and scripting


```{r}
pop_summary <- aggregate(cbind(Height, Fitness, SeedWeight) ~ Population, data = traits, FUN = mean)
write.csv(pop_summary, here::here("results/population_trait_summary.csv"), row.names = FALSE)
```


```{r}
#| eval: false


# exporting plot as a .png
png(here::here("results/height_by_region.png"), width = 800, height = 600)
boxplot(Height ~ Region, data = traits)
dev.off()
```


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Code breakdown: the two save patterns

**Tables.** `write.csv(x, file, row.names = FALSE)` writes a data frame as
CSV. `row.names = FALSE` suppresses the meaningless `"1", "2", "3", ...`
column R would otherwise add — your data already has meaningful identifiers
(`Population`).

**Plots.** Base R plots draw to a *graphics device*. By default that device
is your screen; `png()` opens a file device instead, and everything you plot
lands in the file until `dev.off()` closes it. The sequence is always:

1. `png(path, width, height)` — open the file device (dimensions in pixels)
2. draw: `boxplot(...)`, `hist(...)`, etc.
3. `dev.off()` — close the device and flush the file to disk

Forgetting `dev.off()` is the classic bug: the file stays empty, and every
*next* plot you make goes into it instead of the screen. If plots stop
appearing in your session, run `dev.off()` once and check `results/`.
:::

Put everything into `src/01_explore_traits.R` and run it non-interactively — this is what makes it reproducible and Slurm-friendly:


```{bash}
#| eval: false


Rscript src/01_explore_traits.R
```


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Code breakdown: what makes a script "reproducible"?

A script (rather than typed commands) gives you three things:

1. **A record** — the exact sequence of steps, reviewable and citable.
2. **Re-runnability** — same script + same raw data = same outputs, on your
   laptop or on a Slurm node, forever. This is the working definition of
   reproducibility.
3. **Scale** — only a script can be handed to `sbatch`, scheduled overnight,
   or re-run when the data updates.

The skeleton to aim for in `src/01_explore_traits.R`:

```{r}
#| eval: false
# 01_explore_traits.R — first look at red spruce fitness traits
# Usage: Rscript src/01_explore_traits.R

# --- setup: paths (relative to project root) -----------------------------
raw_file <- here::here("data/raw/red_spruce_fitness_traits.txt")
out_dir  <- here::here("results")

# --- load ----------------------------------------------------------------
traits <- read.delim(raw_file, header = TRUE, sep = "\t",
                     stringsAsFactors = FALSE)

# --- inspect -------------------------------------------------------------
dim(traits)
str(traits)
colSums(is.na(traits))

# --- analyze -------------------------------------------------------------
pop_summary <- aggregate(cbind(Height, Fitness, SeedWeight) ~ Population,
                         data = traits, FUN = mean)

# --- save ----------------------------------------------------------------
write.csv(pop_summary, here::here(out_dir, "population_trait_summary.csv"),
          row.names = FALSE)
```

Notice the section comments — `setup / load / inspect / analyze / save` —
and that not a single absolute path appears anywhere. Anyone can clone the
project, run one command, and get your results.
:::


## Wrap up  
### Idea check


Answer these in your own words before moving to the exercises — write your answers as comments at the top of `src/01_explore_traits.R`.


1. What is the difference between a vector and a data frame in R?
2. Why prefer a relative path like `"data/raw/red_spruce_fitness_traits.txt"` over an absolute path like `/home/your_username/...`?
3. What does `str(traits)` tell you that `head(traits)` does not?
4. Why does treating `Region` as a `factor` (rather than plain character text) matter for `boxplot(Height ~ Region, data = traits)`?
5. If `mean(traits$Height)` returned `NA`, what would you check first, and how would you fix it?
6. Why write outputs into `results/` instead of overwriting anything in `data/raw/`?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Sample answers

1. A vector is one-dimensional and holds a **single type**; a data frame is
   two-dimensional — a list of equal-length vectors, so each column can have
   its own type. Vectors are the building blocks; data frames organize them
   into one-observation-per-row tables.
2. Relative paths travel; absolute paths don't. `data/raw/...` resolves
   correctly wherever the project lives — laptop, cluster, collaborator's
   machine — while `/home/your_username/...` only exists on one computer.
   `here::here()` gives you relative-path portability even when the working
   directory is wrong.
3. `str()` reports the **type of every column** (`chr`/`int`/`num`) and the
   object's shape — information about *all* 326 rows. `head()` only shows the
   first few rows' values; a column mis-parsed as text looks identical in
   `head()` but announces itself in `str()`.
4. A factor carries **fixed, ordered levels**, which the boxplot formula uses
   to define the groups and their axis order. Plain characters usually work
   here too, but factors give you control (alphabetical, or a biologically
   meaningful order), guarantee every category appears even with zero counts,
   and are what modeling functions like `lm()` expect for categorical
   predictors.
5. Check for missing values: `sum(is.na(traits$Height))` — here, 13. Fix with
   `mean(traits$Height, na.rm = TRUE)`. The deeper question — *why* are 13
   heights missing and are they missing at random? — matters before you
   publish any number computed on the remaining 313.
6. `data/raw/` is the **immutable source**: if analysis outputs overwrite it,
   you can no longer distinguish input from product, and re-running the
   script gives different (corrupted) results. Everything in `results/` is
   disposable and regenerable from raw + script; the raw file is not.
:::


### Practical exercises


Work through these in order inside an R session, then consolidate the working code into `src/01_explore_traits.R`.


### Exercise 1 — Load and confirm the shape

Load the dataset and confirm it has 326 rows and 19 columns.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
traits <- read.delim(here::here("data/raw/red_spruce_fitness_traits.txt"),
                     header = TRUE, sep = "\t", stringsAsFactors = FALSE)
dim(traits)
```

`dim()` returns `326 19` — 326 trees, 19 variables. If you had gotten 327 rows,
the likely culprit would be a header misread; 1 row, a wrong separator; and a
single column, tabs not being recognized (check the actual file with
`head data/raw/red_spruce_fitness_traits.txt` in the shell).
:::


### Exercise 2 — Map every column to its type

Run `str(traits)` and list every column name along with its type (`chr`, `int`, `num`).

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

`str(traits)` prints the full map. Grouped:

- `chr`: `Family`, `Population`, `Location`, `Region`
- `int`: `Tree`, `Elevation`
- `num`: `Latitude`, `Longitude`, `SeedWeight`, `Germination`, `Survival`,
  `Height`, `Fitness`, `PC1`, `PC2`, `Family_Homozygosity`,
  `Population_Homozygosity`, `Genetic_Diversity`, `Genetic_Load`

That is 4 text, 2 integer, 13 numeric columns = 19 total. Sanity-check the
*biological* plausibility of the types too: identifiers as text ✓, elevation
in whole meters as integer ✓, proportions like `Survival` as numeric ✓. A
type that surprises you (say, `Height` as `chr`) means some row contains
non-numeric junk — find it with
`traits$Height[is.na(as.numeric(traits$Height))]`-style checks.
:::


### Exercise 3 — Summary statistics, and the NA trap

Compute overall summary statistics (mean, sd, min, max) for `Height` and `Fitness`.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
stats_of <- function(x) {
  c(mean = mean(x, na.rm = TRUE),
    sd   = sd(x, na.rm = TRUE),
    min  = min(x, na.rm = TRUE),
    max  = max(x, na.rm = TRUE))
}

rbind(Height  = stats_of(traits$Height),
      Fitness = stats_of(traits$Fitness))
```

Expected values: Height mean **30.03**, sd **4.64**, range **19.39–47.56**;
Fitness mean **14.01**, sd **7.32**, range **0.00–41.43**.

The critical detail: without `na.rm = TRUE` every one of these returns `NA`,
because `Height` is missing for 13 trees and `Fitness` for 12. Writing your
own `stats_of()` wrapper (a function — see the quick reference) bakes the
`na.rm` in once instead of repeating it eight times. Note how `rbind()`
stacks the two named rows into one table — the same row-binding idea you'll
meet again when combining populations in the population genetics module.
:::


### Exercise 4 — Count trees per group

Use `table()` to count trees per `Population` and per `Region`. Which population has the most trees sampled?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
table(traits$Population)
table(traits$Region)
which.max(table(traits$Population))
```

- `table(traits$Region)` → **C: 178, E: 101, M: 47** — 178 + 101 + 47 = 326,
  so every tree has a region.
- The most-sampled population is **G57 with 11 trees**; most populations have
  4–7 trees, and a couple (`CRR`, `XCS`) have only 2.

`which.max()` returns the *name* of the maximum entry, automating the
eyeball-scan. Keep the small populations in mind for Exercise 6 — means
computed from 2 trees are far noisier than means from 11.
:::


### Exercise 5 — Subsetting with two conditions

Subset the data to trees in `Region == "C"` with `Survival > 0.8`. How many trees meet this criterion?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
c_healthy <- traits[traits$Region == "C" & traits$Survival > 0.8, ]
nrow(c_healthy)
```

**116 trees** (out of 178 in region C). Points to notice:

- The comma inside `[ , ]` is mandatory — without it you select *columns*
  matching a logical vector, a different (and here, nonsense) operation.
- `&` combines the two conditions; both must hold for a row to survive.
- One tree has `NA` for `Survival`, so its comparison is `NA` and the row is
  dropped silently — usually the right behavior, but be able to explain why.
- Sanity check: `nrow(c_healthy)` = 116 ≤ 178 = `table(traits$Region)["C"]`.
  A filtered count larger than its parent group is impossible; catching that
  means your condition logic is broken (probably `|` where you meant `&`).
:::


### Exercise 6 — Grouped means

Use `aggregate()` or `tapply()` to get mean `Height` and mean `Fitness` per `Population`. Identify the top 3 populations by mean `Fitness`.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
pop_means <- aggregate(cbind(Height, Fitness) ~ Population,
                       data = traits, FUN = mean)
head(pop_means[order(-pop_means$Fitness), ], 3)   # top 3 by Fitness
tail(pop_means[order(-pop_means$Fitness), ], 2)   # bottom of the table, for contrast
```

Top 3 by mean fitness: **G18 (24.53), BRU (24.18), BRA (22.91)**; the lowest
are XWS (2.19) and XSK (3.11).

`order(-pop_means$Fitness)` sorts descending — the minus flips the order of
an otherwise ascending sort. Read the results with the caveat from
Exercise 4 in mind: G18 has only 3 trees, so its mean rests on a tiny
sample; its rank could easily change with more sampling. Grouped means are
only as trustworthy as the per-group counts behind them.
:::


### Exercise 7 — Plots, saved to files

Create a histogram of `SeedWeight` and a boxplot of `Height` by `Region`. Save both as PNG files in `results/`.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
#| eval: false
png(here::here("results/seedweight_hist.png"), width = 800, height = 600)
hist(traits$SeedWeight, breaks = 30,
     main = "Seed weight distribution", xlab = "Seed weight")
dev.off()

png(here::here("results/height_by_region.png"), width = 800, height = 600)
boxplot(Height ~ Region, data = traits,
        main = "Height by region", ylab = "Height (cm)")
dev.off()
```

The three-beat pattern each time: `png()` opens the file, the plot call draws
into it, `dev.off()` closes and flushes. Verify the files exist afterwards
with `list.files(here::here("results"))` — an empty file means you forgot
`dev.off()`; plots appearing on screen instead mean `png()` never ran.
:::


### Exercise 8 — Save the summary table

Save your population-level summary table as `results/population_trait_summary.csv`.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
write.csv(pop_summary,
          here::here("results/population_trait_summary.csv"),
          row.names = FALSE)
```

`row.names = FALSE` keeps the file keyed by `Population` alone — the CSV's
first column will be `Population`, not R's meaningless `1, 2, 3, ...` row
numbers. Inspect it from the shell with
`head results/population_trait_summary.csv` to confirm what your future self
(and your collaborators) will actually receive.
:::


### Exercise 9 — The reproducibility test

Move all working code into `src/01_explore_traits.R` and confirm it runs cleanly end-to-end with `Rscript src/01_explore_traits.R`, producing the expected files in `results/`.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution sketch

The pass/fail test is a clean-room re-run:

```{bash}
#| eval: false
# from the project root
rm -rf results                # delete everything R ever produced
Rscript src/01_explore_traits.R
ls -lh results/                # the script must recreate it all
```

If that sequence recreates `population_trait_summary.csv` plus the two PNGs
with no errors and no warnings you can't explain, your module is
reproducible. If it fails, the usual culprits are, in order: a hard-coded
absolute path, a missing `dir.create(results)` before the first write, and
code that depends on objects you created interactively but never put in the
script. The last one is the reason to build scripts from a fresh R session,
not from the session where you explored.
:::


### Homework 


- Compute summary statistics (mean, sd, min, max) for at least **4 traits** (e.g., `Height`, `Fitness`, `SeedWeight`, `Germination`), broken down by both `Population` and `Region`.
- Produce at least **3 plots**: one histogram, one boxplot grouped by `Region`, and one scatterplot relating two continuous variables (e.g., `Elevation` vs. `Fitness`). Save all three to `results/`.
- Write your interpretation as comments at the bottom of the script, answering:
  - Which population has the highest and lowest mean fitness?
  - Is there a visible relationship between elevation and fitness or seed weight? Describe it in one or two sentences.
- **Bonus (optional):** recreate one plot with `ggplot2` instead of base R.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Hints (not full answers)

- **Grouped by Region first** — it's only 3 groups, so eyeball-able:
  mean Height is **C 29.7 / E 28.9 / M 34.0** and mean Fitness
  **C 14.7 / E 10.6 / M 19.8**. Region M — the group with the *lowest*
  median elevation (570 m, vs. 787 for C and 1333 for E) — is taller *and*
  higher-fitness on average. Per-population, extend the
  `aggregate(... ~ Population)` pattern from Exercise 6; for two grouping
  factors at once, try `aggregate(cbind(...) ~ Region, ...)` separately —
  a `Population × Region` cross needs `table(traits$Population, traits$Region)`
  or `interaction()`.
- **The scatterplot answer you should arrive at:** elevation vs. fitness is a
  **moderate negative relationship** (correlation ≈ −0.36) — higher sites,
  lower mean fitness; elevation vs. seed weight is **weakly positive**
  (≈ +0.25). Both are visible-but-loose trends in the point cloud, not
  tight lines. Remember the two cautions from the plotting section before
  you write the interpretation: association ≠ causation, and region
  structure is confounded with elevation here.
- **ggplot2 bonus:** the same boxplot is
  `ggplot(traits, aes(x = Region, y = Height)) + geom_boxplot()` after
  `install.packages("ggplot2")` and `library(ggplot2)` — the `+` layers
  replace the single-call-with-arguments style of base plotting.
:::
