---
title: "_R_: Population Genetics Foundations"
description: "Allele frequencies, heterozygosity, and a first genotype–environment association"
author:
  - name: "Anoob Prakash"
    email: "prakas89@purdue.edu"
    orcid: "0000-0002-7207-934X"
    url: "https://www.linkedin.com/in/anoob-prakash/"
date: "10/07/2026"
categories:
  - hpc
  - r
  - foundations
  - github
  - popgen


format:
  html:
    toc: true
    toc-location: left
    toc-title: "On this page"
    toc-depth: 3
    page-layout: full
    lightbox: true
    code-fold: false
    code-overflow: wrap
    code-copy: true
    code-tools: false
    code-block-border-left: "#583C88"
    code-block-border-left-size: 5px
    code-block-border-left-style: solid


---


# PART I 
## Goal


Apply your R skills to population-genetics concepts: allele frequency, heterozygosity, genetic diversity, and a first genotype/phenotype–environment association (GEA), using the genetic-parameter columns already present in the red-spruce dataset (`PC1`, `PC2`, `Family_Homozygosity`, `Population_Homozygosity`, `Genetic_Diversity`, `Genetic_Load`).


We don't have a raw genotype (VCF) file for this dataset, so this module teaches the underlying math with a small **simulated** SNP matrix, then applies the same reasoning to the **real, pre-computed** genetic-parameter columns already in your data.


### Prerequisites
::: {.column-body}
_1. Data required data from github_ 



::: {.callout-important collapse="true"}  
## Download data
**Download the raw GitHub file directly into** `data/raw/`


**Copy the code below and run it inside the** `cluster-exercise` **directory**          



```{r}
#| eval: false
dir.create("data/raw", recursive = TRUE, showWarnings = FALSE)
download.file(
"https://raw.githubusercontent.com/anoobvinu07/Genomic_assisted_selection/master/data/FitnessTraits_GeneticParameters_RedSpruce.txt",
      destfile = "data/raw/red_spruce_fitness_traits.txt")
```  


The `-L` flag tells `curl` to follow redirects, and `-o` specifies the local output file. Verify the download before doing any analysis:


Recreate the rest of the project layout if you don't already have it:


```{r}
#| eval: false
dir.create("results", showWarnings = FALSE)
dir.create("src", showWarnings = FALSE)
```
:::


::: {.callout-important collapse="true"}
## Project directory setup
**This excercise assumes the project folder follows the following structure:** 


<pre style="color: #F8F9FA; background-color: #22252A; padding: 1rem; border-radius: 6px;"><code>cluster-exercise/
 ├── data/
 │   └── raw/       # downloaded, unchanged source data
 ├── logs/          # Slurm standard output and error logs
 ├── results/       # analysis products created by R
 └── src/           # R and Slurm scripts</code></pre>
:::


_2. Install the packages used here:_


```{r}
#| eval: false


  install.packages(c("vegan", "here"))
  # Optional, for working with real genotype/VCF files beyond this module:
  # install.packages("BiocManager")
  # BiocManager::install(c("vcfR", "adegenet"))
```


```{r}
#| echo: false
#| eval: false
setwd(dir = "/Users/anoobprakash/Documents/GitHub/anoobvinu07.github.io")
```


:::


```{r}
#| include: false


library(here)
```






## Learning objectives


By the end of this module you should be able to:


- Explain what allele frequency, observed heterozygosity, and expected heterozygosity mean
- Represent genotypes as a simple numeric matrix (0/1/2 coding) and compute allele frequencies from it
- Compute observed vs. expected (Hardy-Weinberg) heterozygosity per locus and interpret the difference
- Run a PCA on genetic markers with `prcomp()` and interpret PC1/PC2 in a population-genetics context
- Test a simple genotype/phenotype–environment association using correlation and linear models
- Recognize the step from a simple correlation to a multivariate method (RDA) used in real genotype-environment association work
- Know when to move from a plain R matrix to dedicated packages (`vcfR`, `adegenet`) for real VCF-scale data


### R for bioinformatics quick reference


| Function/package | Purpose | Example |
|---|---|---|
| `matrix()` | Build a genotype matrix | `matrix(..., nrow, ncol)` |
| `sample()` | Randomly draw values, optionally with replacement and specified probabilities | `sample(0:2, 300, replace = TRUE, prob = c(0.4, 0.4, 0.2))` |
| `rownames`/ `colnames` | Get or assign labels for matrix rows and columns  | `rownames(geno) <- paste0("ind_", 1:n_ind)` |
| `paste0` | Convert inputs to text and concatenate them with no separator  | `paste0("ind",1:30)` |
| `colMeans()` | Column-wise means (used for allele freq.) | `colMeans(geno) / 2` |
| `table()` | Count occurrences of each value | `table(pop_1$genotype)` |
| `prop.table()` | Convert counts to proportions | `prop.table(table(pop_1$genotype))` |
| `rbind()` | Stack data frames row-wise | `rbind(pop_1, pop_2)` |
| `prcomp()` | Principal component analysis | `prcomp(geno, scale. = TRUE)` |
| `cor.test()` | Test correlation between two variables | `cor.test(x, y)` |
| `lm()` | Fit a linear model | `lm(y ~ x1 + x2, data = df)` |
| `vegan::rda()` | Redundancy analysis (multivariate GEA) | `rda(Y ~ x1 + x2, data = df)` |
| `vcfR::read.vcfR()` | Read a real VCF file (beyond this module) | `read.vcfR("file.vcf")` |
| `adegenet::genind` | Genotype container object (beyond this module) | `df2genind(...)` |



## Basics of population genetics via simulation  
### A minimal SNP genotype matrix in R


Genotypes at a biallelic SNP are usually coded as the count of the alternate allele: `0` (homozygous reference), `1` (heterozygous), `2` (homozygous alternate). Let's simulate a toy dataset — 30 individuals, 10 loci:


```{r}
set.seed(123)
n_ind  <- 30 # number of individuals
n_loci <- 10 # number of SNP loci
```
This code creates a simulated genotype matrix where `rows are individuals`, `columns are SNP loci`, and each cell contains a `diploid genotype coded as 0, 1, or 2`.


```{r}
geno <- matrix(
  sample(0:2,                # sample values from 0, 1 or 2
  n_ind * n_loci,            # no. of draws 30 x 10 or size of the matrix
  replace = TRUE,            # each genotype category can be drawn repeatedly
  prob = c(0.4, 0.4, 0.2)),  # sampling probability for genotypes 0,1 and 2
  nrow = n_ind,              # number of rows - filled with ind IDs
  ncol = n_loci              # number of columns - filled with SNP IDs
)
```


::: {.callout-note icon="false" appearance="simple" collapse="true" }  


# Logic check  
::: {.panel-tabset}


## Genotype coding  
For a biallelic SNP, the number represents the count of the alternate allele carried by an individual:  


| Value | Genotypes | Biological meaning | 
|---|---|---| 
| 0 | 0/0 | Homozygous reference; zero alternate alleles |
| 1 | 0/1 | Heterozygous; one alternate allele |
| 2 | 1/1 | Homozygous alternate; two alternate alleles |  


So, for example, a `1` in row `ind_7` and column `locus_3` means `individual 7` is heterozygous at `SNP locus 3`. 


## Genotype probability


Set the probabilty of each genotype for the simlutaion run with `prob` argument of `sample()` function. 


| Genotype value | Sampling probability | 
|---|---|---| 
| 0 | 0.40 |
| 1 | 0.40 |
| 2 | 0.20 |


:::





 
:::


```{r}
rownames(geno) <- paste0("ind_", 1:n_ind)
colnames(geno) <- paste0("locus_", 1:n_loci)


geno[1:5, 1:5]
```


### Allele frequencies


Allele frequency is the `proportion of all gene copies at a locus that are a particular allele`. 
$$p + q =1$$ 


Each genotype value is the count of the alternate allele out of 2, so the mean genotype divided by 2 is the alternate allele frequency for that locus:


```{r}
alt_freq <- colMeans(geno) / 2
ref_freq <- 1 - alt_freq


data.frame(locus = names(alt_freq),       # col 1 filled with locus names
           alt_freq = round(alt_freq, 3), # col 2 filled with alt freq
           ref_freq = round(ref_freq, 3)) # col 3 is filled with ref freq
```


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: allele frequency by hand

Suppose one locus has 20 genotyped individuals: **12 are `0/0`, 5 are `0/1`, and 3 are `1/1`**.

Each individual carries 2 allele copies, so the denominator is $2N = 40$ gene copies:

- Alternate allele copies: $(0 \times 12) + (1 \times 5) + (2 \times 3) = 11$
- Reference allele copies: $(2 \times 12) + (1 \times 5) + (0 \times 3) = 29$

$$p_{alt} = \frac{11}{40} = 0.275 \qquad p_{ref} = \frac{29}{40} = 0.725 \qquad p_{alt} + p_{ref} = 1$$

The shortcut `mean(geno_column) / 2` works for exactly the same reason: the mean of a
0/1/2 column is the average number of alternate alleles **per individual**, so dividing
by 2 converts it to a per-gene-copy frequency.
:::


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: why divide by 2?

The mean of a 0/1/2 genotype column counts alleles **per individual**, but allele
frequency counts alleles **per gene copy**. A diploid individual has 2 copies, hence
the division by 2. Sanity checks that always hold:

- `alt_freq` must lie between 0 and 1
- `alt_freq + ref_freq` must equal exactly 1
- a column that is entirely `0` or entirely `2` is **monomorphic** (no variation) and
  carries no information for a GEA — real pipelines filter these out along with
  low-frequency variants
:::



# PART II   


## Goals 
In this exercise, you will simulate genotypes for two populations at one biallelic locus, calculate observed allele and genotype frequencies, compare the populations, and combine them into a single dataset.  


### Learning objectives
By the end of this section of the module, you should be able to:  


- Use `sample()` to simulate diploid genotypes (`AA`, `Aa`, and `aa`).   
- Calculate genotype frequencies from observed genotype counts.  
- Calculate allele frequencies \(p\) and \(q\) from genotype counts.  
- Compare populations with different allele frequencies and population sizes.  
- Combine population-level data with `rbind()`.  
- Explain why combining structured populations can change apparent genotype frequencies, even when each population was generated under Hardy–Weinberg expectations.




## Simulating two populations at one locus  


Lets start off by creating two populations with of same size (n=20), but different allele A frequency. This exercise simulates genotypes at one biallelic locus in two populations. You will use sample() to assign genotypes, calculate genotype frequencies with table() and prop.table(), then combine populations with rbind(). 


For this example, the two alleles are A and a, with three possible diploid genotypes:


| Genotype | Meaning |
| --- | --- |
| _AA_ | Two copies of allele _A_ | 
| _Aa_ | One copy of _A_ and one copy of _a_ | 
| _aa_ | Two copies of allele _a_ | 


At a biallelic locus:


$$p + q = 1 $$
where,  
    _p_ is the frequency of allele _A_ and;  
    _q_ is the frequency of allele _a_.  



Under _Hardy–Weinberg_ expectations, genotype probabilities are:


$$AA = p^2$$
$$Aa = 2pq$$


$$aa = q^2$$
 


::: {.callout-important icon="false" appearance="default" collapse="true"}
## Concept review : H-W equilibrium  


**[What is Hardy-Weinberg equilibrium?](https://bio.libretexts.org/Bookshelves/Introductory_and_General_Biology/General_Biology_(Boundless)/19%3A_The_Evolution_of_Populations/19.01%3A_Population_Evolution/19.1C%3A_Hardy-Weinberg_Principle_of_Equilibrium)**


The key expectataion under Hardy-Weinberg equilibrium is $p^2$ for _AA_, $2pq$ for _Aa_, and $q^2$ for _aa_, where $p + q = 1$. In finite populations, observed values vary around thse expectations because of random sampling, variation becomes more obvious at small sample sizes. 
:::


The code below uses those probabilities to generate individuals in each population. Because genotypes are randomly sampled, observed frequencies will usually be close to but not identical to the expected frequencies.


```{r}
set.seed(42)


# Population 1: 20 individuals; allele A frequency p = 0.80
p_1 <- 0.80  
q_1 <- 1 - p_1


pop_1 <- data.frame(population = "Pop_1",
                    # sample function for choosing genotypes
                    genotype = sample(c("AA", "Aa", "aa"), # possible genotype values
                      size = 20,                           # no. of individuals to simulate
                      replace = TRUE,                      # genotypes can  occur more than once
                      # p2 + 2pq + q2
                      prob = c(p_1^2, 2 * p_1 * q_1, q_1^2)) # probability of each genotype
                      )


head(pop_1) # peek into the pop_1 dataframe


# Population 2: 20 individuals; allele A frequency p = 0.30
p_2 <- 0.30
q_2 <- 1 - p_2


pop_2 <- data.frame(population = "Pop_2",
                    genotype = sample(c("AA", "Aa", "aa"),
                      size = 20,
                      replace = TRUE,
                      prob = c(p_2^2, 2 * p_2 * q_2, q_2^2))
                      )


head(pop_2) # peek into the pop_2 dataframe
```


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Code breakdown  


  - sample() randomly assigns each individual one of three possible genotypes: AA, Aa, or aa.    
  - replace = TRUE allows genotype categories to occur repeatedly.    
  - prob = c($p^2$, $2.p.q$, $q^2$) supplies the Hardy–Weinberg genotype probabilities.      


:::


For population 1, where _p_ = 0.80 and _q_ = 0.20, the expected genotype frequencies are:
$$AA = 0.80^2 = 0.64$$
$$Aa = 2 \times 0.80 \times 0.20 = 0.32$$  
$$aa = 0.20^2 = 0.04$$
Therefore, Population 1 should countain mostly _AA_ individuals.


For Population 2, where _p_ = 0.30 and _q_ = 0.70, the expected genotype frequencies are:
$$AA = 0.30^2 = 0.09$$
$$Aa = 2 \times 0.30 \times 0.70 = 0.42$$  
$$aa = 0.70^2 = 0.49$$
Therefore, population2 should contain mostly _aa_ individuals. 



### Calculate genotype frequencies   
Use table() to count each genotype and prop.table() to convert those counts into frequencies.   
```{r}
# Count observed genotypes in each population
geno_count_pop1 <- table(pop_1$genotype)
geno_count_pop2 <- table(pop_2$genotype)



# Convert genotype counts to genotype frequencies  
geno_prop_pop1 <- prop.table(table(pop_1$genotype))
geno_prop_pop2 <- prop.table(table(pop_2$genotype))


# Combine individuals from the two populations
pooled_population <- rbind(pop_1, pop_2)


geno_count_pooled <- table(pooled_population$genotype)
geno_prop_pooled <- prop.table(table(pooled_population$genotype))



# Build one comparison table
geno_summary <- data.frame(
  genotype = c("AA", "Aa", "aa"),
  count_pop1 = as.integer(geno_count_pop1[c("AA", "Aa", "aa")]),
  freq_pop1  = as.numeric(geno_prop_pop1[c("AA", "Aa", "aa")]),
  count_pop2 = as.integer(geno_count_pop2[c("AA", "Aa", "aa")]),
  freq_pop2  = as.numeric(geno_prop_pop2[c("AA", "Aa", "aa")]),
  count_pooled = as.integer(geno_count_pooled[c("AA", "Aa", "aa")]),
  freq_pooled  = as.numeric(geno_prop_pooled[c("AA", "Aa", "aa")])
)


geno_summary


# The genotype frequencies should sum up to 1. 
sum(geno_prop_pop1)
sum(geno_prop_pop2)
sum(geno_prop_pooled)
```


::: {.callout-note icon="false" appearance="default" collapse="true"}
## Code breakdown  
   
  - rbind() or "row" bind stacks the two population data frames row by row.  
  - table() counts each observed genotype.    
  
:::


### Calculate allele frequencies  
Genotype frequencies describe the proportion of individuals in each genotype class. Allele frequencies instead count all gene copies. Each diploid individual carries two allele copies:  
 
| Genotype | Number of _A_ alleles | Number of _a_ alleles |
| --- | --- | --- |
|_AA_ | 2 | 0 |
|_Aa_ | 1 | 1 | 
| _aa_ | 0 | 2 | 



$$p_{A} = \frac{2n_{AA} + n_{Aa}}{2N} $$


$$q_{a} = \frac{2n_{aa} + n_{Aa}}{2N} $$
Where each _AA_ individuals contributes two _A_ alleles, each _aa_ individual contributes two _aa_ alleles, and each heterozygote contributes one copy of each allele.


To calculate the allele frequency for Population 1:


```{r}
# Count each genotype in Population 1
counts_1 <- geno_count_pop1


# Number of individuals
N_1 <- nrow(pop_1)


# Count A and a allele copies
n_A_1 <- 2 * counts_1["AA"] + counts_1["Aa"]
n_a_1 <- 2 * counts_1["aa"] + counts_1["Aa"]


# Convert allele counts to allele frequencies
p_A_1 <- n_A_1 / (2 * N_1)
q_a_1 <- n_a_1 / (2 * N_1)


p_A_1
q_a_1
p_A_1 + q_a_1
```


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: the numbers behind Population 1

With `set.seed(42)`, Population 1 has **9 `AA`, 10 `Aa`, 1 `aa`**. Plugging those
counts into the formulas:

$$p_{A} = \frac{(2 \times 9) + 10}{40} = \frac{28}{40} = 0.70$$

$$q_{a} = \frac{(2 \times 1) + 10}{40} = \frac{12}{40} = 0.30$$

Note that $p_A + q_a = 1$ exactly — this is not a coincidence but a bookkeeping
identity: every one of the $2N = 40$ gene copies is counted exactly once on either
the numerator side of $p_A$ or of $q_a$.

Two things worth noticing:

1. The **observed** allele frequency (0.70) differs from the **simulation** input
   (0.80) — random sampling in a finite population (N = 20) shifts the estimate.
   Smaller samples drift further from the input value.
2. The observed genotype counts (9/10/1) do not match the H–W expectations for
   $p = 0.70$ (12.25/6.00/1.75 out of 20) — here we observe an **excess of
   heterozygotes**. With only 20 individuals, deviations like this are expected
   sampling noise, not biology.
:::


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: expected vs. observed counts

Expected counts are just $N \times$ expected frequency. For Population 1
($p = 0.80$, $N = 20$):

| Genotype | Expected frequency | Expected count (of 20) | Observed count (seed 42) |
|---|---|---|---|
| AA | 0.64 | 12.8 | 9 |
| Aa | 0.32 | 6.4 | 10 |
| aa | 0.04 | 0.8 | 1 |

The expected `aa` count is **less than 1 individual**. This is why formal
goodness-of-fit tests (like `chisq.test()`) are unreliable here: the standard rule
of thumb requires expected counts of at least ~5 per category. With $q^2 = 0.04$ and
N = 20, you would need roughly $5 / 0.04 = 125$ individuals just to expect 5 `aa`
individuals. Rare genotype classes need large samples.
:::



## Excercises 


### Exercise 1 — Allele frequencies for Population 2

Estimate the allele frequencies $p_A$ and $q_a$ for **Population 2**, following the
same steps used for Population 1: count genotypes, count allele copies, and divide
by $2N$. Verify that $p + q = 1$.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
counts_2 <- geno_count_pop2
N_2 <- nrow(pop_2)

n_A_2 <- 2 * counts_2["AA"] + counts_2["Aa"]
n_a_2 <- 2 * counts_2["aa"] + counts_2["Aa"]

p_A_2 <- unname(as.numeric(n_A_2 / (2 * N_2)))
q_a_2 <- unname(as.numeric(n_a_2 / (2 * N_2)))

p_A_2
q_a_2
p_A_2 + q_a_2
```

With `set.seed(42)`, Population 2 has **2 `AA`, 10 `Aa`, 8 `aa`**, so:

$$p_{A} = \frac{(2 \times 2) + 10}{40} = \frac{14}{40} = 0.35$$

$$q_{a} = \frac{(2 \times 8) + 10}{40} = \frac{26}{40} = 0.65$$

Again the observed $p$ (0.35) sits above the simulation input (0.30) — finite-sample
noise. The two populations are strongly differentiated at this locus:
$p_A$ = 0.70 vs. 0.35.
:::


### Exercise 2 — Allele frequencies for the pooled population

Compute the allele frequencies for the **pooled** population (Pop_1 + Pop_2).
Compare them to the within-population values, and explain the pattern: where does
the pooled estimate fall, and why?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
counts_p <- geno_count_pooled
N_p <- nrow(pooled_population)

n_A_p <- 2 * counts_p["AA"] + counts_p["Aa"]
n_a_p <- 2 * counts_p["aa"] + counts_p["Aa"]

p_A_p <- unname(as.numeric(n_A_p / (2 * N_p)))
q_a_p <- unname(as.numeric(n_a_p / (2 * N_p)))

p_A_p
q_a_p
```

Pooled counts are **11 `AA`, 20 `Aa`, 9 `aa`** (40 individuals, 80 gene copies):

$$p_{A} = \frac{(2 \times 11) + 20}{80} = \frac{42}{80} = 0.525$$

$$q_{a} = \frac{(2 \times 9) + 20}{80} = \frac{38}{80} = 0.475$$

The pooled $p_A$ = 0.525 is almost exactly the **average** of the two
within-population frequencies:

$$\frac{p_{A,1} + p_{A,2}}{2} = \frac{0.70 + 0.35}{2} = 0.525$$

(With unequal population sizes, the pooled frequency is a weighted average, weighted
by each population's size.) Mixing does not create or destroy alleles — it just
averages them. But as the next exercise shows, it **does** distort *genotype*
frequencies.
:::


### Exercise 3 — Pooling changes genotype frequencies

Using the `geno_summary` table you built earlier:

1. Compute the **expected** genotype frequencies for the pooled population from its
   allele frequencies ($p^2$, $2pq$, $q^2$).
2. Compare them to the **observed** pooled genotype frequencies.
3. Compute the expected genotype frequencies you would get by averaging the two
   populations' *own* H–W expectations.
4. Which comparison do the observed pooled frequencies resemble more? Why?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
# 1. H-W expectations from the POOLED allele frequencies
exp_pooled_from_pool_p <- c(AA = p_A_p^2, Aa = 2 * p_A_p * q_a_p, aa = q_a_p^2)

# 2. Observed pooled genotype frequencies
obs_pooled <- as.numeric(geno_prop_pooled[c("AA", "Aa", "aa")])

# 3. Average of the two populations' OWN H-W expectations
exp_avg_of_pops <- c(
  AA = (p_1^2 + p_2^2) / 2,
  Aa = (2 * p_1 * q_1 + 2 * p_2 * q_2) / 2,
  aa = (q_1^2 + q_2^2) / 2
)

rbind(
  observed            = obs_pooled,
  HW_from_pooled_p    = exp_pooled_from_pool_p,
  avg_of_pop_HW       = exp_avg_of_pops
)
```

Using the seed-42 numbers ($p_{pool} = 0.525$):

- H–W from pooled $p$: $AA = 0.276$, $Aa = 0.499$, $aa = 0.226$
- Average of the two populations' own expectations: $AA = 0.365$, $Aa = 0.370$, $aa = 0.265$

Here is the interesting part: at this seed, the **observed** pooled frequencies
(0.275 / 0.500 / 0.225) land almost exactly on the H–W line from the pooled $p$ —
the Wahlund deficit is invisible. Compare the two reference rows to see why the
deficit *should* be there: H–W from the pooled $p$ predicts 49.9% heterozygotes,
but heterozygotes in the mixture can only come from the two populations' own
heterozygote proportions, which average 37%. The systematic expectation is a
**13-point heterozygote deficit** — the Wahlund effect.

Why can't we see it? At $N = 20$ per population, the standard error of a genotype
proportion near 0.5 is about $\sqrt{0.25/20} \approx 0.11$ — the same size as the
deficit itself. In this seed, Pop_1 happened to draw a heterozygote *excess*
(10 observed vs. 6.4 expected), which cancels the Wahlund gap almost perfectly.

The takeaway is not that the effect is fake, but that **at small N it is buried in
noise**. Pooling two H–W populations systematically produces a heterozygote deficit
relative to H–W expectations computed from the pooled allele frequencies — no
population evolved, only the *reference frame* changed. Part III re-runs this exact
comparison at $N = 500$ per population, where the deficit stands out unmistakably.
:::


### Exercise 4 — Allele frequencies from the genotype matrix

Return to the 30 individuals $\times$ 10 loci matrix `geno` from Part I.

1. Compute the alternate allele frequency for every locus **by hand** using allele
   counting (`2*n2 + n1` divided by `2*N`), not `colMeans()`.
2. Verify that your result matches `colMeans(geno) / 2` exactly.
3. Compute the **minor allele frequency (MAF)** for each locus. Which locus is
   closest to fixation? Which is most variable?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
N <- nrow(geno)

# 1. Allele counting, locus by locus
n_alt <- 2 * colSums(geno == 2) + colSums(geno == 1)
alt_freq_manual <- n_alt / (2 * N)

# 2. Cross-check with the shortcut
all.equal(as.numeric(alt_freq_manual), as.numeric(alt_freq))

# 3. Minor allele frequency
maf <- pmin(alt_freq, 1 - alt_freq)
sort(round(maf, 3))
```

`all.equal()` confirms the two computations agree — `colMeans(geno)/2` is just
algebraically compressed allele counting: the mean of a 0/1/2 column is
$(n_{het} + 2 \cdot n_{homAlt}) / N$, which divided by 2 is exactly the alternate
allele frequency.

For the MAF, the least variable locus is `locus_6` (alt = 0.317, MAF $\approx$ 0.317)
and the most variable is `locus_10` (alt = 0.450, MAF $\approx$ 0.450). The most
informative loci for GEA are the ones with MAF near 0.5, which is why filtering
pipelines typically drop loci with MAF below 0.05: rare variants carry little signal
and inflate noise.
:::


### Exercise 5 — Explore the simulation

Change **one thing at a time** and re-run the whole Part II pipeline. Before you run
each variant, write down what you expect to happen; then check.

- **Sample size:** set both populations to `size = 500`. Do observed frequencies land
  closer to the H–W expectations?
- **Similar populations:** set $p_2 = 0.75$. What happens to the pooled genotype
  frequencies and the Wahlund gap?
- **Divergent populations:** set $p_2 = 0.05$. What happens?
- **Different seed:** change `set.seed(42)` to another number. Which quantities
  change (counts, frequencies) and which relationships stay fixed
  ($p + q = 1$, pooled $p$ = average of within-population $p$)?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution — what you should see

**Sample size (N = 500):** observed genotype frequencies converge on the H–W
expectations. Sampling noise shrinks like $1/\sqrt{N}$ — ten times the sample, about
three times less noise. This is the single most reliable way to distinguish
"sampling noise" from "real deviation from H–W".

**Similar populations ($p_2 = 0.75$):** the two populations are nearly
indistinguishable at this locus. Pooling barely changes anything — there is almost no
Wahlund effect because there is almost no structure. Weak structure = weak
heterozygote deficit.

**Divergent populations ($p_2 = 0.05$):** the contrast is extreme — Pop_1 is nearly
all `AA`, Pop_2 nearly all `aa`. The pooled sample looks like a classic
heterozygote-deficient population even though each source population is in perfect
H–W equilibrium. Strong structure = strong apparent deficit. This is the situation
that most often fools naive analyses of mixed samples (e.g., pooling before
computing heterozygosity in a VCF).

**Different seed:** the counts and frequencies all shift slightly, but the
*relationships* hold: $p + q = 1$ in every population and the pooled $p$ always
equals the (size-weighted) average of within-population $p$. Identities survive
randomness; estimates don't.

One practical gotcha when you change seeds or parameters: if a genotype class
happens not to occur (e.g., no `aa` individuals in Pop_1), `table()` silently drops
that level and indexing `counts["aa"]` returns `NA`. Keep all three classes with
`table(factor(pop_1$genotype, levels = c("AA", "Aa", "aa")))` so counts of zero
are explicit.
:::



# PART III 


### Observed and expected heterozygosity


**Observed heterozygosity ($H_{o}$):** the proportion of individuals with genotype `1` (heterozygous) at a locus.


**Expected heterozygosity ($H_{e}$):** what you'd expect under Hardy-Weinberg equilibrium, given the allele frequencies: $H_{e} = 2pq$, where _p_ and _q_ are the reference and alternate allele frequencies.


**Concept review: [What is Hardy-Weinberg equilibrium?](https://bio.libretexts.org/Bookshelves/Introductory_and_General_Biology/General_Biology_(Boundless)/19%3A_The_Evolution_of_Populations/19.01%3A_Population_Evolution/19.1C%3A_Hardy-Weinberg_Principle_of_Equilibrium)**


```{r}
Ho <- colMeans(geno == 1)
He <- 2 * ref_freq * alt_freq


het_table <- data.frame(locus = names(Ho), 
                        Ho = round(Ho, 3), 
                        He = round(He, 3),
                        diff = round(Ho - He, 3))
het_table
```


If `Ho` is consistently lower than `He` across loci, that's a signature of inbreeding or population substructure — the same underlying idea behind the `Family_Homozygosity` and `Population_Homozygosity` columns already computed in the [real world dataset](https://github.com/anoobvinu07/Genomic_assisted_selection/blob/master/data/FitnessTraits_GeneticParameters_RedSpruce.txt) used in the previous modules.


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: heterozygosity by hand

Take the hand-calculation locus from Part I: **12 `0/0`, 5 `0/1`, 3 `1/1`**
($N = 20$, $p_{alt} = 0.275$, $p_{ref} = 0.725$).

**Observed heterozygosity** — just count the heterozygotes:

$$H_{o} = \frac{n_{het}}{N} = \frac{5}{20} = 0.25$$

**Expected heterozygosity** — the H–W heterozygote frequency at those allele
frequencies:

$$H_{e} = 2pq = 2 \times 0.725 \times 0.275 = 0.399$$

This locus has fewer heterozygotes than expected ($H_o < H_e$). The standard way to
summarize the gap is the **inbreeding coefficient**:

$$F = 1 - \frac{H_{o}}{H_{e}} = 1 - \frac{0.25}{0.399} = 0.373$$

- $F = 0$: the population matches H–W exactly at this locus
- $F > 0$: heterozygote **deficit** (inbreeding, substructure, or chance)
- $F < 0$: heterozygote **excess** (e.g., overdominance, or sampling noise)

Notice how the pieces from Part II assemble here: **allele counts → allele
frequencies → H–W expectations → observed vs. expected → F**. Every downstream
population-genetics quantity is built from the same two inputs, $p$ and $N$.
:::


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: $H_e$ is maximized at $p = 0.5$

$H_e = 2pq$ is a parabola in $p$ — it peaks when both alleles are equally common:

| $p$ | $q$ | $H_e = 2pq$ |
|---|---|---|
| 0.1 | 0.9 | 0.18 |
| 0.3 | 0.7 | 0.42 |
| 0.5 | 0.5 | **0.50** |
| 0.7 | 0.3 | 0.42 |
| 0.9 | 0.1 | 0.18 |
| 1.0 | 0.0 | 0.00 |

Three consequences worth internalizing:

1. **Near-fixed loci are uninformative.** At $p = 0.95$, $H_e = 0.095$ — almost no
   variation to work with. This is why MAF filtering exists (Exercise 4, Part II).
2. **Heterozygosity measures evenness, not richness.** A population fixed for a
   different allele at every locus and one fixed for the same allele at every locus
   both have $H_e = 0$.
3. **Homozygosity and diversity are two views of the same quantity**: for a biallelic
   locus, $H_e + \text{homozygosity} = 1$. In the red-spruce data, this is why
   `Population_Homozygosity` and `Genetic_Diversity` should be near-perfect mirror
   images of each other (negatively correlated) — check it in the practical exercises.
:::


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: reading the simulated het_table

For the Part I simulation (seed 123), the table shows:

- Mean $H_o \approx 0.403$, mean $H_e \approx 0.473$, so $H_o < H_e$ at **most** loci,
  and the overall $F \approx 0.147$.
- The largest deficit is at `locus_1` ($H_o = 0.267$ vs. $H_e = 0.480$).
- The largest excess is at `locus_10` ($H_o = 0.567$ vs. $H_e = 0.495$).

Here is the subtle part — this deficit is **built into the simulation**, not a
biological process. We drew *genotypes* with probabilities `c(0.4, 0.4, 0.2)` rather
than drawing *alleles* and pairing them randomly. Those genotype probabilities imply
an allele frequency of $p = (0 \times 0.4 + 1 \times 0.4 + 2 \times 0.2)/2 = 0.4$,
and H–W at that frequency would give $H_e = 2(0.4)(0.6) = 0.48$ — but our draw only
produces heterozygotes 40% of the time. So the simulation has an inbreeding
coefficient of about $F = 1 - 0.40/0.48 \approx 0.17$ by construction.

The lesson: **always ask how a simulation generates its data before interpreting the
output.** A genotype-frequency simulator and an allele-frequency simulator with the
same $p$ produce different heterozygosities — which is precisely the $H_o$ vs $H_e$
distinction.
:::


### The Wahlund effect: connecting Part II to Part III

In Part II you saw that pooling two populations distorts genotype frequencies. Here
is the same demonstration at a sample size where the pattern is unambiguous. This
time, simulate two large populations (N = 500 each) at one locus:

```{r}
set.seed(7)
N_w <- 500

# Same structure as Part II, just bigger
pop_big_1 <- sample(c("AA", "Aa", "aa"), N_w, replace = TRUE,
                    prob = c(p_1^2, 2 * p_1 * q_1, q_1^2))   # p = 0.80
pop_big_2 <- sample(c("AA", "Aa", "aa"), N_w, replace = TRUE,
                    prob = c(p_2^2, 2 * p_2 * q_2, q_2^2))   # p = 0.30

pool_big <- c(pop_big_1, pop_big_2)

Ho_of <- function(g) sum(g == "Aa") / length(g)
p_of  <- function(g) (2 * sum(g == "AA") + sum(g == "Aa")) / (2 * length(g))

data.frame(
  group = c("Pop_1", "Pop_2", "Pooled"),
  p_A = round(c(p_of(pop_big_1), p_of(pop_big_2), p_of(pool_big)), 3),
  Ho = round(c(Ho_of(pop_big_1), Ho_of(pop_big_2), Ho_of(pool_big)), 3),
  He_from_p = round(2 * c(p_of(pop_big_1), p_of(pop_big_2), p_of(pool_big)) *
                       (1 - c(p_of(pop_big_1), p_of(pop_big_2), p_of(pool_big))), 3)
)
```

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Reading the Wahlund table

With seed 7 you should see numbers like these:

| Group | $p_A$ | $H_o$ | $H_e$ (from its own $p$) |
|---|---|---|---|
| Pop_1 | 0.800 | 0.332 | 0.320 |
| Pop_2 | 0.307 | 0.418 | 0.426 |
| **Pooled** | **0.553** | **0.375** | **0.494** |

Read the last row carefully — both comparisons matter:

1. **Within each population, $H_o \approx H_e$.** Each population is in H–W
   equilibrium, as designed. No inbreeding, no structure — nothing is wrong with
   either population.
2. **The pooled row shows $H_o = 0.375 \ll H_e = 0.494$**, an apparent
   heterozygote deficit with $F \approx 0.24$ — **out of nowhere**, if you didn't
   know the sample was structured.
3. The pooled $H_e$ (0.494) is also **higher than either population's own $H_e$**
   (0.320, 0.426). Pooling mixes allele frequencies toward the middle ($p = 0.553$,
   near 0.5 where $H_e$ peaks), so the *expected* diversity of the mixture exceeds
   the diversity of any real component.

This is the **Wahlund effect**: combining structured populations inflates apparent
expected heterozygosity and produces an apparent heterozygote deficit — mimicking
inbreeding even when every subpopulation is perfectly outbred.

Why does this matter for your work? If you compute heterozygosity from a mixed VCF
without accounting for population structure, substructure masquerades as
inbreeding. It also runs the other way: the PCA section below is exactly the tool you
use to *detect* the structure that causes Wahlund effects — the connection between
these two sections is the heart of "everything connects."

In the red-spruce dataset, `Population_Homozygosity` computed across structured
regions contains this same mixing effect, which is one reason it is analyzed
**by region** and alongside `PC1`/`PC2`.
:::


### Exercises: heterozygosity

### Exercise 6 — Ho and He for the two populations

Using `pop_1` and `pop_2` from Part II (N = 20 each):

1. Compute $H_o$ (observed frequency of `Aa`) and $H_e$ (from the observed $p$) for
   each population.
2. Compute $H_o$ and $H_e$ for the **pooled** population.
3. Is the pooled $H_o$ below the pooled $H_e$? Is the deficit larger or smaller than
   in the N = 500 demonstration above? Why?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
Ho_pop1 <- sum(pop_1$genotype == "Aa") / nrow(pop_1)
Ho_pop2 <- sum(pop_2$genotype == "Aa") / nrow(pop_2)
Ho_pool <- sum(pooled_population$genotype == "Aa") / nrow(pooled_population)

He_pop1 <- 2 * p_A_1 * q_a_1
He_pop2 <- 2 * p_A_2 * q_a_2
He_pool <- 2 * p_A_p * q_a_p

data.frame(
  group = c("Pop_1", "Pop_2", "Pooled"),
  Ho = round(c(Ho_pop1, Ho_pop2, Ho_pool), 3),
  He = round(c(He_pop1, He_pop2, He_pool), 3),
  F = round(1 - c(Ho_pop1, Ho_pop2, Ho_pool) /
              c(He_pop1, He_pop2, He_pool), 3)
)
```

With seed 42, all three groups happen to show $H_o \approx 0.50$ (and the within-pop
$H_e$ values are 0.420 and 0.455), so the Wahlund deficit is essentially invisible:
$F_{pool} \approx -0.003$ — a hair of heterozygote *excess* — even though the
populations are just as differentiated as in the big simulation.

The reason is **noise swamping signal**: with N = 20, the standard error of a
proportion near 0.5 is $\sqrt{0.25/20} \approx 0.11$, while the expected Wahlund
deficit for this configuration is $H_{e,pool} - \overline{H_{o,within}} \approx
0.49 - 0.37 \approx 0.12$ — the same order as the noise. With N = 500, the standard
error shrinks to $\approx 0.022$ and the deficit stands out clearly.

Practical takeaway: heterozygote deficits (and excesses) in small samples are
uninterpretable without either larger N or replicated loci. This is why
genome-scale analyses average over thousands of loci — the per-locus noise cancels
while a systematic deficit (structure, inbreeding) persists.
:::


### Exercise 7 — Average across loci

For the `geno` matrix from Part I:

1. Add a per-locus inbreeding coefficient $F = 1 - H_o/H_e$ to `het_table`.
2. Which single locus departs most from H–W in *each* direction?
3. Compute the mean $H_o$, mean $H_e$, and multi-locus $F$ across all 10 loci. Why is
   averaging across loci more trustworthy than any single locus?

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution

```{r}
# Compute F from the unrounded Ho and He, then attach it to the table
het_table$F <- round(1 - Ho / He, 3)
het_table

mean(het_table$Ho)
mean(het_table$He)
1 - mean(het_table$Ho) / mean(het_table$He)
```

With seed 123: the strongest deficit is `locus_1` ($F = 0.44$), the strongest
excess is `locus_10` ($F = -0.15$), and the multi-locus value is
$F \approx 0.147$.

Averaging across loci works because each locus is an independent draw of the same
underlying process. Random per-locus deviations — which push individual $F$ values
both up and down — cancel out, while a systematic process (inbreeding, structure, or
in our case the simulation's genotype probabilities) shifts all loci in the same
direction and survives the average. This is the same logic as Exercise 6 writ large:
more data, same signal, less noise. It is also exactly why real genomic studies
report genome-wide average $H_o/H_e$ rather than single-locus values.
:::


### Exercise 8 — Connect to the real data

Using the red-spruce dataset:

1. Plot `Genetic_Diversity` against `Population_Homozygosity`.
2. Fit `lm(Population_Homozygosity ~ Genetic_Diversity, data = traits)`.
3. Interpret the slope in light of the identity $H_e + \text{homozygosity} = 1$
   from the worked example above.

::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Solution sketch

```{r}
#| eval: false
plot(traits$Genetic_Diversity, traits$Population_Homozygosity,
     pch = 19, xlab = "Genetic diversity", ylab = "Population homozygosity")
summary(lm(Population_Homozygosity ~ Genetic_Diversity, data = traits))
```

For a single biallelic locus, expected homozygosity is $p^2 + q^2 = 1 - 2pq = 1 -
H_e$, so the two quantities are deterministically complementary. If these columns
were computed from the same set of loci and the same weighting, you should find a
strongly **negative** relationship — though not necessarily a slope of exactly
$-1$, since the columns may use different locus sets, weightings, or estimators.
Scatter around a strong negative trend is expected with real, noisy, structured
data; deviations can also reflect population structure (Wahlund again) inflating
homozygosity in mixed samples. Check the outlier populations against their `Region`
on the PCA plot.
:::


### PCA for population structure


On the simulated matrix:


```{r}
pca_sim <- prcomp(geno, scale. = TRUE)
plot(pca_sim$x[, 1], pca_sim$x[, 2],
     xlab = "PC1", ylab = "PC2", main = "Simulated genotype PCA")
```


Now look at the **real** PC1/PC2 already computed for the red-spruce trees (from actual genetic markers, not simulated):


```{r}
#| eval: false
traits <- read.table("data/raw/red_spruce_fitness_traits.txt", header = TRUE, sep = "\t")
```


```{r}
#| echo: false
traits_file <- here::here(
  "data",
  "raw",
  "red_spruce_fitness_traits.txt"
)


stopifnot(file.exists(traits_file))


traits <- read.delim(
  traits_file,
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = FALSE
)
```
```{r}
traits$Region <- factor(traits$Region)
plot(traits$PC1, traits$PC2,
     col = as.integer(traits$Region), pch = 19,
     xlab = "PC1", ylab = "PC2", main = "Genetic PCA colored by region")
legend("topright", legend = levels(traits$Region),
       col = 1:nlevels(traits$Region), pch = 19)
```


In population genetics, PC1/PC2 from a genetic marker PCA usually capture the strongest axes of population structure — often correlated with geography (isolation by distance) or major environmental gradients.


::: {.callout-tip icon="false" appearance="default" collapse="true"}
## Worked example: what a PCA of the two-population simulation would show

Go back to the Wahlund simulation (two populations, $p = 0.8$ vs. $p = 0.3$, N = 500
each) and imagine running `prcomp()` on a 1000 $\times$ 1 matrix of genotypes.
Because there is only one locus, the "PCA" is trivial — but if you extend the
simulation to hundreds of loci that all differ between the populations in the same
direction (like locally adapted loci), PC1 would capture exactly one thing: **which
population an individual belongs to**.

That is what PC1/PC2 do in real data. When the red-spruce PCA shows points colored by
`Region` separating along PC1, you are looking at the multilocus version of the
allele-frequency difference $p_{A,1} = 0.70$ vs. $p_{A,2} = 0.35$ from Part II. The
PCA doesn't create information — it summarizes thousands of per-locus frequency
differences into two axes you can plot.

The loop closes here: **structure (Part II pooling) → heterozygote deficit (Wahlund)
→ PCA detects that structure → and the GEA/RDA step must account for it**, because a
climate variable correlated with region would otherwise look like adaptation when it
is really just ancestry. This is why `PC1` sits in the red-spruce dataset alongside
the climate variables.
:::



## How everything connects

Everything in this module is one pipeline applied at increasing scale. Trace it
end-to-end with a single question: **"why is this locus heterozygote-deficient?"**

| Step | Concept | Where you computed it | Red-spruce equivalent |
|---|---|---|---|
| 1 | Count genotypes | `table(pop_1$genotype)` | — |
| 2 | Allele frequencies $p, q$ | `colMeans(geno)/2`, allele counting | — |
| 3 | H–W expectations $p^2, 2pq, q^2$ | `prob = c(p^2, 2pq, q^2)` | — |
| 4 | Observed vs. expected heterozygosity | `Ho`, `He`, `het_table` | `Genetic_Diversity` |
| 5 | The gap: $F = 1 - H_o/H_e$ | Exercise 7 | `Family_Homozygosity` (conceptually related) |
| 6 | Structure causes the gap | Wahlund effect (N = 500 demo) | `Region`, `PC1`, `PC2` |
| 7 | Structure detected multivariately | `prcomp()` PCA plot | `PC1`, `PC2` columns |
| 8 | Structure + environment | `cor.test()`, `lm()` | `Elevation`, `Latitude` |
| 9 | Many loci + environment jointly | `vegan::rda()` (stretch) | RDA / genomic offset |

Three threads run through the whole table:

- **The same two quantities everywhere.** Every row is computed from genotype counts
  and sample size. Heterozygosity, F, the Wahlund effect, PCA loadings, and even the
  RDA ordination are all functions of allele frequencies — the only thing that
  changes is how many loci and how many environmental axes you let in.
- **Structure is the recurring confounder.** Pooling populations inflates apparent
  diversity (Part II, Wahlund), heterozygote deficits can mean either inbreeding or
  structure (Part III), and the PCA that reveals structure (`PC1`/`PC2`) is also the
  covariate you must control before interpreting any genotype–environment
  association (Part III's `lm()` and `rda()`). A climate correlation that survives
  after accounting for structure is evidence of local adaptation; one that
  disappears was ancestry all along.
- **Scale changes confidence, not logic.** With N = 20 at one locus, noise dominates
  (Exercise 6). With N = 500, the Wahlund effect is obvious. With 10 loci, averaging
  helps (Exercise 7). With genome-wide SNPs, per-locus estimates are noise but
  multivariate summaries (PCA, RDA) become powerful. The math you did by hand on a
  2-allele, 20-individual toy is the same math inside `adegenet`, `LEA`, and your
  genomic-offset pipelines.

When you reach the stretch exercise — `rda()` with `Genetic_Diversity`,
`Genetic_Load`, `PC1`, `PC2` as responses and `Elevation`, `Latitude` as predictors —
notice what it is: rows 8 and 9 of the table. A correlation tests one
response–predictor pair; the RDA tests all of them jointly, in the same way the PCA
summarized all loci jointly. Multivariate = the same idea, more axes.


## Roadmap beyond this module


When you're ready to work with real genotype data instead of pre-computed summaries:


- `vcfR::read.vcfR()` reads a VCF file into R
- `adegenet::df2genind()` / `vcfR::vcfR2genind()` convert genotypes into population-genetics-aware objects with built-in allele-frequency, heterozygosity, and Fst functions
- `LEA` (Bioconductor) and `vegan::rda()` scale the GEA approach above to genome-wide SNP data


None of this is required for this module's exercises — it's here so you know where basics you learn connects to your actual genomic analysis pipelines.


## Wrap Up
### Idea check


Answer these before the practical exercises.


1. What does an alternate allele frequency of 0.3 at a locus mean biologically?
2. Why can observed heterozygosity (Ho) be lower than expected heterozygosity (He) in a real population?
3. What do you think `Population_Homozygosity` is measuring, and how would you expect it to relate to `Genetic_Diversity` — positively or negatively?
4. In a PCA built from genetic markers, what do PC1 and PC2 typically represent in a population-genetics context?
5. Why test `cor.test(Genetic_Diversity, Elevation)` rather than just assuming a relationship exists because both vary by population?
6. What is the conceptual difference between running a separate `lm()` for each genetic variable versus running one `rda()` with all of them as a response matrix?


### Practical exercises


1. Build the simulated genotype matrix (30 individuals × 10 loci) exactly as shown, and compute the alternate allele frequency for every locus.
2. Compute observed and expected heterozygosity for each locus and identify which locus shows the largest Ho − He gap.
3. Run `prcomp()` on the simulated genotype matrix and plot PC1 vs. PC2.
4. Load the real red-spruce dataset and plot the real `PC1` vs. `PC2`, colored by `Region`. Does the pattern look like discrete clusters or a continuous gradient?
5. Run `cor.test()` between `Population_Homozygosity` and `Elevation`, and separately between `Genetic_Diversity` and `Latitude`. Report the correlation coefficient, p-value, and a one-sentence interpretation for each.
6. Fit `lm(Genetic_Load ~ Elevation + Latitude, data = traits)` and interpret the two coefficients — which predictor has a stronger association, and in which direction?
7. **Stretch exercise:** run the `vegan::rda()` example above using `Genetic_Diversity`, `Genetic_Load`, `PC1`, `PC2` as responses and `Elevation`, `Latitude` as predictors. Produce the ordination biplot and describe what you see in 2–3 sentences.
8. Save every summary table (allele frequencies, heterozygosity table, correlation results) and every plot into `results/`.

Solutions for the simulation-based items (1–3) are worked through in Part I, the
Part II exercises, and Exercise 7 — attempt them before unfolding those callouts.


### Homework 


Build `src/02_pop_genetics_intro.R`, combining the simulated-data fundamentals with a real-data GEA-style analysis.


**Requirements:**


- Part A (fundamentals): allele frequency and Ho/He table for the simulated genotype matrix, plus the PCA plot.
- Part B (applied): using the real dataset, test associations between **at least two** genetic-parameter columns (e.g., `Population_Homozygosity`, `Genetic_Diversity`, `Genetic_Load`) and **at least two** environmental predictors (`Elevation`, `Latitude`, `Longitude`), using correlation and/or `lm()`.
- A written interpretation (half a page, as comments or a `results/README.md`) answering: Do the genetic parameters show a spatial pattern with latitude or elevation? What ecological or demographic process might explain this (e.g., isolation by distance, local adaptation, drift in small/marginal populations)?
- **Bonus (optional):** complete the `vegan::rda()` ordination and include the biplot with a short interpretation of which environmental axis drives more separation among genetic variables.


**Submission format:** `src/02_pop_genetics_intro.R`, all generated outputs in `results/`, and the written interpretation.
