---
title: "R for Bioinformatics: Population Genetics Foundations"
description: "allele frequencies, heterozygosity, and a first genotype–environment association"
author:
  - name: "Anoob Prakash"
    email: "prakas89@purdue.edu"
    orcid: "0000-0002-7207-934X"
    url: "https://www.linkedin.com/in/anoob-prakash/"
date: "9/11/2026"
categories:
  - hpc
  - r
  - bioinformatics
  - github
  - popgen

format:
  html:
    toc: true
    toc-location: left
    toc-title: "On this page"
    toc-depth: 3
    page-layout: full
    lightbox: true
    code-fold: false
    code-overflow: wrap
    code-copy: true
    code-tools: false
    code-block-border-left: "#583C88"
    code-block-border-left-size: 5px
    code-block-border-left-style: solid

---

# PART I 
## Goal

Apply the R skills from Module 1 to population-genetics concepts: allele frequency, heterozygosity, genetic diversity, and a first genotype/phenotype–environment association (GEA), using the genetic-parameter columns already present in the red-spruce dataset (`PC1`, `PC2`, `Family_Homozygosity`, `Population_Homozygosity`, `Genetic_Diversity`, `Genetic_Load`).

We don't have a raw genotype (VCF) file for this dataset, so this module teaches the underlying math with a small **simulated** SNP matrix, then applies the same reasoning to the **real, pre-computed** genetic-parameter columns already in your data.

### Prerequisites
::: {.column-body}
_1. Data required data from github_ 


::: {.callout-important collapse="true"}  
## Download data
**Download the raw GitHub file directly into** `data/raw/`

**Copy the code below and run it inside the** `cluster-exercise` **directory**          


```{r}
#| eval: false
dir.create("data/raw", recursive = TRUE, showWarnings = FALSE)
download.file(
"https://raw.githubusercontent.com/anoobvinu07/Genomic_assisted_selection/master/data/FitnessTraits_GeneticParameters_RedSpruce.txt",
      destfile = "data/raw/red_spruce_fitness_traits.txt")
```  

The `-L` flag tells `curl` to follow redirects, and `-o` specifies the local output file. Verify the download before doing any analysis:

Recreate the rest of the project layout if you don't already have it:

```{r}
#| eval: false
dir.create("results", showWarnings = FALSE)
dir.create("src", showWarnings = FALSE)
```
:::

::: {.callout-important collapse="true"}
## Project directory setup
**This excercise assumes the project folder follows the following structure:** 

<pre style="color: #F8F9FA; background-color: #22252A; padding: 1rem; border-radius: 6px;"><code>cluster-exercise/
 ├── data/
 │   └── raw/       # downloaded, unchanged source data
 ├── logs/          # Slurm standard output and error logs
 ├── results/       # analysis products created by R
 └── src/           # R and Slurm scripts</code></pre>
:::

_2. Install the packages used here:_

```{r}
#| eval: false

  install.packages(c("vegan"))
  # Optional, for working with real genotype/VCF files beyond this module:
  # install.packages("BiocManager")
  # BiocManager::install(c("vcfR", "adegenet"))
```

```{r}
#| echo: false
#| eval: false
setwd(dir = "/Users/anoobprakash/Documents/GitHub/anoobvinu07.github.io")
```

:::

```{r}
#| include: false

library(here)
```





## Learning objectives

By the end of this module you should be able to:

- Explain what allele frequency, observed heterozygosity, and expected heterozygosity mean
- Represent genotypes as a simple numeric matrix (0/1/2 coding) and compute allele frequencies from it
- Compute observed vs. expected (Hardy-Weinberg) heterozygosity per locus and interpret the difference
- Run a PCA on genetic markers with `prcomp()` and interpret PC1/PC2 in a population-genetics context
- Test a simple genotype/phenotype–environment association using correlation and linear models
- Recognize the step from a simple correlation to a multivariate method (RDA) used in real genotype-environment association work
- Know when to move from a plain R matrix to dedicated packages (`vcfR`, `adegenet`) for real VCF-scale data

### R for bioinformatics quick reference

| Function/package | Purpose | Example |
|---|---|---|
| `matrix()` | Build a genotype matrix | `matrix(..., nrow, ncol)` |
| `sample()` | Randomly draw values, optionally with replacement and specified probabilities | `sample(0:2, 300, replace = TRUE, prob = c(0.4, 0.4, 0.2))` |
| `rownames`/ `colnames` | Get or assign labels for matrix rows and columns  | `rownames(geno) <- paste0("ind_", 1:n_ind)` |
| `paste0` | Convert inputs to text and concatenate them with no separator  | `paste0("ind",1:30)` |
| `colMeans()` | Column-wise means (used for allele freq.) | `colMeans(geno) / 2` |
| `prcomp()` | Principal component analysis | `prcomp(geno, scale. = TRUE)` |
| `cor.test()` | Test correlation between two variables | `cor.test(x, y)` |
| `lm()` | Fit a linear model | `lm(y ~ x1 + x2, data = df)` |
| `vegan::rda()` | Redundancy analysis (multivariate GEA) | `rda(Y ~ x1 + x2, data = df)` |
| `vcfR::read.vcfR()` | Read a real VCF file (beyond this module) | `read.vcfR("file.vcf")` |
| `adegenet::genind` | Genotype container object (beyond this module) | `df2genind(...)` |


## Basics of population genetics via simulation  
### A minimal SNP genotype matrix in R

Genotypes at a biallelic SNP are usually coded as the count of the alternate allele: `0` (homozygous reference), `1` (heterozygous), `2` (homozygous alternate). Let's simulate a toy dataset — 30 individuals, 10 loci:

```{r}
set.seed(123)
n_ind  <- 30 # number of individuals
n_loci <- 10 # number of SNP loci
```
This code creates a simulated genotype matrix where `rows are individuals`, `columns are SNP loci`, and each cell contains a `diploid genotype coded as 0, 1, or 2`.

```{r}
geno <- matrix(
  sample(0:2,                # sample values from 0, 1 or 2
  n_ind * n_loci,            # no. of draws 30 x 10 or size of the matrix
  replace = TRUE,            # each genotype category can be drawn repeatedly
  prob = c(0.4, 0.4, 0.2)),  # sampling probability for genotypes 0,1 and 2
  nrow = n_ind,              # number of rows - filled with ind IDs
  ncol = n_loci              # number of columns - filled with SNP IDs
)
```

::: {.callout-note icon="false" appearance="simple" collapse="true" }  

# Logic check  
::: {.panel-tabset}

## Genotype coding  
For a biallelic SNP, the number represents the count of the alternate allele carried by an individual:  

| Value | Genotypes | Biological meaning | 
|---|---|---| 
| 0 | 0/0 | Homozygous reference; zero alternate alleles |
| 1 | 0/1 | Heterozygous; one alternate allele |
| 2 | 1/1 | Homozygous alternate; two alternate alleles |  

So, for example, a `1` in row `ind_7` and column `locus_3` means `individual 7` is heterozygous at `SNP locus 3`. 

## Genotype probability

Set the probabilty of each genotype for the simlutaion run with `prob` argument of `sample()` function. 

| Genotype value | Sampling probability | 
|---|---|---| 
| 0 | 0.40 |
| 1 | 0.40 |
| 2 | 0.20 |

:::




 
:::

```{r}
rownames(geno) <- paste0("ind_", 1:n_ind)
colnames(geno) <- paste0("locus_", 1:n_loci)

geno[1:5, 1:5]
```

### Allele frequencies

Allele frequency is the `proportion of all gene copies at a locus that are a particular allele`. 
$$p + q =1$$ 

Each genotype value is the count of the alternate allele out of 2, so the mean genotype divided by 2 is the alternate allele frequency for that locus:

```{r}
alt_freq <- colMeans(geno) / 2
ref_freq <- 1 - alt_freq

data.frame(locus = names(alt_freq),       # col 1 filled with locus names
           alt_freq = round(alt_freq, 3), # col 2 filled with alt freq
           ref_freq = round(ref_freq, 3)) # col 3 is filled with ref freq
```


# PART II   

## Goals 
In this exercise, you will simulate genotypes for two populations at one biallelic locus, calculate observed allele and genotype frequencies, compare the populations, and combine them into a single dataset.  

### Learning objectives
By the end of this section of the module, you should be able to:  

- Use `sample()` to simulate diploid genotypes (`AA`, `Aa`, and `aa`).   
- Calculate genotype frequencies from observed genotype counts.  
- Calculate allele frequencies \(p\) and \(q\) from genotype counts.  
- Compare populations with different allele frequencies and population sizes.  
- Combine population-level data with `rbind()`.  
- Explain why combining structured populations can change apparent genotype frequencies, even when each population was generated under Hardy–Weinberg expectations.



## Simulating two populations at one locus  

Lets start off by creating two populations with of same size (n=20), but different allele A frequency. This exercise simulates genotypes at one biallelic locus in two populations. You will use sample() to assign genotypes, calculate genotype frequencies with table() and prop.table(), then combine populations with rbind(). 

For this example, the two alleles are A and a, with three possible diploid genotypes:

| Genotype | Meaning |
| --- | --- |
| _AA_ | Two copies of allele _A_ | 
| _Aa_ | One copy of _A_ and one copy of _a_ | 
| _aa_ | Two copies of allele _a_ | 

At a biallelic locus:

$$p + q = 1 $$
where,  
    _p_ is the frequency of allele _A_ and;  
    _q_ is the frequency of allele _a_.  


Under _Hardy–Weinberg_ expectations, genotype probabilities are:

$$AA = p^2$$
$$Aa = 2pq$$

$$aa = q^2$$
 

::: {.callout-important icon="false" appearance="default" collapse="true"}
## Concept review : H-W equilibrium  

**[What is Hardy-Weinberg equilibrium?](https://bio.libretexts.org/Bookshelves/Introductory_and_General_Biology/General_Biology_(Boundless)/19%3A_The_Evolution_of_Populations/19.01%3A_Population_Evolution/19.1C%3A_Hardy-Weinberg_Principle_of_Equilibrium)**

The key expectataion under Hardy-Weinberg equilibrium is $p^2$ for _AA_, $2pq$ for _Aa_, and $q^2$ for _aa_, where $p + q = 1$. In finite populations, observed values vary around thse expectations because of random sampling, variation becomes more obvious at small sample sizes. 
:::

The code below uses those probabilities to generate individuals in each population. Because genotypes are randomly sampled, observed frequencies will usually be close to but not identical to the expected frequencies.

```{r}
set.seed(42)

# Population 1: 20 individuals; allele A frequency p = 0.80
p_1 <- 0.80  
q_1 <- 1 - p_1

pop_1 <- data.frame(population = "Pop_1",
                    # sample function for choosing genotypes
                    genotype = sample(c("AA", "Aa", "aa"), # possible genotype values
                      size = 20,                           # no. of individuals to simulate
                      replace = TRUE,                      # genotypes can  occur more than once
                      # p2 + 2pq + q2
                      prob = c(p_1^2, 2 * p_1 * q_1, q_1^2)) # probability of each genotype
                      )

head(pop_1) # peek into the pop_1 dataframe

# Population 2: 20 individuals; allele A frequency p = 0.30
p_2 <- 0.30
q_2 <- 1 - p_2

pop_2 <- data.frame(population = "Pop_2",
                    genotype = sample(c("AA", "Aa", "aa"),
                      size = 20,
                      replace = TRUE,
                      prob = c(p_2^2, 2 * p_2 * q_2, q_2^2))
                      )

head(pop_2) # peek into the pop_2 dataframe
```

::: {.callout-note icon="false" appearance="default" collapse="true"}
## Code breakdown  

  - sample() randomly assigns each individual one of three possible genotypes: AA, Aa, or aa.    
  - replace = TRUE allows genotype categories to occur repeatedly.    
  - prob = c($p^2$, $2.p.q$, $q^2$) supplies the Hardy–Weinberg genotype probabilities.      

:::

For population 1, where _p_ = 0.80 and _q_ = 0.20, the expected genotype frequencies are:
$$AA = 0.80^2 = 0.64$$
$$Aa = 2 \times 0.80 \times 0.20 = 0.32$$  
$$aa = 0.20^2 = 0.04$$
Therefore, Population 1 should countain mostly _AA_ individuals.

For Population 2, where _p_ = 0.30 and _q_ = 0.70, the expected genotype frequencies are:
$$AA = 0.30^2 = 0.09$$
$$Aa = 2 \times 0.30 \times 0.70 = 0.42$$  
$$aa = 0.70^2 = 0.49$$
Therefore, population2 should contain mostly _aa_ individuals. 


### Calculate genotype frequencies   
Use table() to count each genotype and prop.table() to convert those counts into frequencies.   
```{r}
# Count observed genotypes in each population
geno_count_pop1 <- table(pop_1$genotype)
geno_count_pop2 <- table(pop_2$genotype)


# Convert genotype counts to genotype frequencies  
geno_prop_pop1 <- prop.table(table(pop_1$genotype))
geno_prop_pop2 <- prop.table(table(pop_2$genotype))

# Combine individuals from the two populations
pooled_population <- rbind(pop_1, pop_2)

geno_count_pooled <- table(pooled_population$genotype)
geno_prop_pooled <- prop.table(table(pooled_population$genotype))


# Build one comparison table
geno_summary <- data.frame(
  genotype = c("AA", "Aa", "aa"),
  count_pop1 = as.integer(geno_count_pop1[c("AA", "Aa", "aa")]),
  freq_pop1  = as.numeric(geno_prop_pop1[c("AA", "Aa", "aa")]),
  count_pop2 = as.integer(geno_count_pop2[c("AA", "Aa", "aa")]),
  freq_pop2  = as.numeric(geno_prop_pop2[c("AA", "Aa", "aa")]),
  count_pooled = as.integer(geno_count_pooled[c("AA", "Aa", "aa")]),
  freq_pooled  = as.numeric(geno_prop_pooled[c("AA", "Aa", "aa")])
)

geno_summary

# The genotype frequencies should sum up to 1. 
sum(geno_prop_pop1)
sum(geno_prop_pop2)
sum(geno_prop_pooled)
```

::: {.callout-note icon="false" appearance="default" collapse="true"}
## Code breakdown  
   
  - rbind() or "row" bind stacks the two population data frames row by row.  
  - table() counts each observed genotype.    
  
:::

### Calculate allele frequencies  
Genotype frequencies describe the proportion of individuals in each genotype class. Allele frequencies instead count all gene copies. Each diploid individual carries two allele copies:  
 
| Genotype | Number of _A_ alleles | Number of _a_ alleles |
| --- | --- | --- |
|_AA_ | 2 | 0 |
|_Aa_ | 1 | 1 | 
| _aa_ | 0 | 2 | 


$$p_{A} = \frac{2n_{AA} + n_{Aa}}{2N} $$

$$q_{a} = \frac{2n_{aa} + n_{Aa}}{2N} $$
Where each _AA_ individuals contributes two _A_ alleles, each _aa_ individual contributes two _aa_ alleles, and each heterozygote contributes one copy of each allele.

To calculate the allele frequency for Population 1:

```{r}
# Count each genotype in Population 1
counts_1 <- geno_count_pop1

# Number of individuals
N_1 <- nrow(pop_1)

# Count A and a allele copies
n_A_1 <- 2 * counts_1["AA"] + counts_1["Aa"]
n_a_1 <- 2 * counts_1["aa"] + counts_1["Aa"]

# Convert allele counts to allele frequencies
p_A_1 <- n_A_1 / (2 * N_1)
q_a_1 <- n_a_1 / (2 * N_1)

p_A_1
q_a_1
p_A_1 + q_a_1
```


## Excercises 
- **Estimate allele frequencies for Population 2**

- **Combine the pooled population data with the Population 1 and Population 2 data**  
  - What changed for the pooled data? What could be the reason?   

- **Try changing the simulations**  
  - Change one value at a time: change population size, make populations genetically similar, make populations drastically different, change the seed (set.seed(?))  








# PART III 

### Observed and expected heterozygosity

**Observed heterozygosity ($H_{o}$):** the proportion of individuals with genotype `1` (heterozygous) at a locus.

**Expected heterozygosity ($H_{e}$):** what you'd expect under Hardy-Weinberg equilibrium, given the allele frequencies: $H_{e} = 2pq$, where _p_ and _q_ are the reference and alternate allele frequencies.

**Concept review: [What is Hardy-Weinberg equilibrium?](https://bio.libretexts.org/Bookshelves/Introductory_and_General_Biology/General_Biology_(Boundless)/19%3A_The_Evolution_of_Populations/19.01%3A_Population_Evolution/19.1C%3A_Hardy-Weinberg_Principle_of_Equilibrium)**

```{r}
Ho <- colMeans(geno == 1)
He <- 2 * ref_freq * alt_freq

het_table <- data.frame(locus = names(Ho), 
                        Ho = round(Ho, 3), 
                        He = round(He, 3),
                        diff = round(Ho - He, 3))
het_table
```

If `Ho` is consistently lower than `He` across loci, that's a signature of inbreeding or population substructure — the same underlying idea behind the `Family_Homozygosity` and `Population_Homozygosity` columns already computed in the [real world dataset](https://github.com/anoobvinu07/Genomic_assisted_selection/blob/master/data/FitnessTraits_GeneticParameters_RedSpruce.txt) used in the previous modules.

### PCA for population structure

On the simulated matrix:

```{r}
pca_sim <- prcomp(geno, scale. = TRUE)
plot(pca_sim$x[, 1], pca_sim$x[, 2],
     xlab = "PC1", ylab = "PC2", main = "Simulated genotype PCA")
```

Now look at the **real** PC1/PC2 already computed for the red-spruce trees (from actual genetic markers, not simulated):

```{r}
#| eval: false
traits <- read.table("data/raw/red_spruce_fitness_traits.txt", header = TRUE, sep = "\t")
```

```{r}
#| echo: false
traits_file <- here::here(
  "data",
  "raw",
  "red_spruce_fitness_traits.txt"
)

stopifnot(file.exists(traits_file))

traits <- read.delim(
  traits_file,
  header = TRUE,
  stringsAsFactors = FALSE,
  check.names = FALSE
)
```
```{r}
traits$Region <- factor(traits$Region)
plot(traits$PC1, traits$PC2,
     col = as.integer(traits$Region), pch = 19,
     xlab = "PC1", ylab = "PC2", main = "Genetic PCA colored by region")
legend("topright", legend = levels(traits$Region),
       col = 1:nlevels(traits$Region), pch = 19)
```

In population genetics, PC1/PC2 from a genetic marker PCA usually capture the strongest axes of population structure — often correlated with geography (isolation by distance) or major environmental gradients.





## Roadmap beyond this module

When you're ready to work with real genotype data instead of pre-computed summaries:

- `vcfR::read.vcfR()` reads a VCF file into R
- `adegenet::df2genind()` / `vcfR::vcfR2genind()` convert genotypes into population-genetics-aware objects with built-in allele-frequency, heterozygosity, and Fst functions
- `LEA` (Bioconductor) and `vegan::rda()` scale the GEA approach above to genome-wide SNP data

None of this is required for this module's exercises — it's here so you know where basics you learn connects to your actual genomic analysis pipelines.

## Wrap Up
### Idea check

Answer these before the practical exercises.

1. What does an alternate allele frequency of 0.3 at a locus mean biologically?
2. Why can observed heterozygosity (Ho) be lower than expected heterozygosity (He) in a real population?
3. What do you think `Population_Homozygosity` is measuring, and how would you expect it to relate to `Genetic_Diversity` — positively or negatively?
4. In a PCA built from genetic markers, what do PC1 and PC2 typically represent in a population-genetics context?
5. Why test `cor.test(Genetic_Diversity, Elevation)` rather than just assuming a relationship exists because both vary by population?
6. What is the conceptual difference between running a separate `lm()` for each genetic variable versus running one `rda()` with all of them as a response matrix?

### Practical exercises

1. Build the simulated genotype matrix (30 individuals × 10 loci) exactly as shown, and compute the alternate allele frequency for every locus.
2. Compute observed and expected heterozygosity for each locus and identify which locus shows the largest Ho − He gap.
3. Run `prcomp()` on the simulated genotype matrix and plot PC1 vs. PC2.
4. Load the real red-spruce dataset and plot the real `PC1` vs. `PC2`, colored by `Region`. Does the pattern look like discrete clusters or a continuous gradient?
5. Run `cor.test()` between `Population_Homozygosity` and `Elevation`, and separately between `Genetic_Diversity` and `Latitude`. Report the correlation coefficient, p-value, and a one-sentence interpretation for each.
6. Fit `lm(Genetic_Load ~ Elevation + Latitude, data = traits)` and interpret the two coefficients — which predictor has a stronger association, and in which direction?
7. **Stretch exercise:** run the `vegan::rda()` example above using `Genetic_Diversity`, `Genetic_Load`, `PC1`, `PC2` as responses and `Elevation`, `Latitude` as predictors. Produce the ordination biplot and describe what you see in 2–3 sentences.
8. Save every summary table (allele frequencies, heterozygosity table, correlation results) and every plot into `results/`.

### Homework 

Build `src/02_pop_genetics_intro.R`, combining the simulated-data fundamentals with a real-data GEA-style analysis.

**Requirements:**

- Part A (fundamentals): allele frequency and Ho/He table for the simulated genotype matrix, plus the PCA plot.
- Part B (applied): using the real dataset, test associations between **at least two** genetic-parameter columns (e.g., `Population_Homozygosity`, `Genetic_Diversity`, `Genetic_Load`) and **at least two** environmental predictors (`Elevation`, `Latitude`, `Longitude`), using correlation and/or `lm()`.
- A written interpretation (half a page, as comments or a `results/README.md`) answering: Do the genetic parameters show a spatial pattern with latitude or elevation? What ecological or demographic process might explain this (e.g., isolation by distance, local adaptation, drift in small/marginal populations)?
- **Bonus (optional):** complete the `vegan::rda()` ordination and include the biplot with a short interpretation of which environmental axis drives more separation among genetic variables.

**Submission format:** `src/02_pop_genetics_intro.R`, all generated outputs in `results/`, and the written interpretation.



